---
title: "Preparing Data for DataStream"
editor: 
  markdown: 
    wrap: sentence
---

After cleaning and reshaping a dataset, we need to verify that the data conforms to DataStream's data schema.

By the end of this chapter, you should be able to:

-   Check that a water quality dataset follows the DataStream data schema.
-   Diagnose specific schema violations for a water quality dataset.
-   Perform final QA/QC checks on a dataset to check for common DataStream formatting issues.
-   Create a validation report and a QA/QC report to share with team members.

This chapter is longer than most because it is designed as a reference workflow in R. You do not need to memorize every function or code chunk. Focus on understanding the overall sequence and return to it as needed throughout your project. The broad steps of the chapter are to:

1. Download and import the DataStream schema into R
2. Extract field rules and allowed values from the schema
3. Create a water quality dataset
4. Run schema validation checks on the water quality dataset
5. Run final QA/QC checks for common issues

<br>

## The DataStream Data Schema

DataStream uses a standardized data format based on the WQX (Water Quality Exchange) standard for the Exchange of Water Quality Data, which was originally developed by the United States Environmental Protection Agency and the United States Geological Survey.
This allows different groups to share results in a common format, and increases a dataset's **Interoperability** and **Reusability** (i.e., the I and R in FAIR principles).

The original, raw data that we receive will almost certainly not already adhere to the DataStream data schema.
A data owner may provide multiple CSV files or spreadsheets with different column names, inconsistent site names, wide-format measurements, non-standard units, or values that do not match DataStream's controlled vocabularies.

In a reproducible data rescue workflow, we want to use R to transform the original files into a DataStream-formatted table.
We then export that table as a CSV, which the relevant data owner will then upload to DataStream.

### Upload Templates

DataStream provides Excel and Google Sheets upload templates to help users understand how their data should be structured for upload.
These templates include field names, examples, allowed values, and lookup tables.
They are especially helpful for visualizing the expected DataStream format and for understanding how different fields relate to one another.

As you work through this chapter, you should open the current DataStream upload template and use it as a visual reference: DataStream Upload Template (link)

The template contains several important sheets:

| Sheet                       | Purpose                                                                                                                          |
|-------------------------|-----------------------------------------------|
| `Allowed Values-Valeurs`    | Lists DataStream fields, whether they are required, optional, or conditional, their field types, examples, and allowed values.   |
| `CharacteristicName LOOKUP` | Provides additional guidance for `CharacteristicName` values, including fields that may be required for certain characteristics. |
| `ResultUnit LOOKUP`         | Provides information about valid `ResultUnit` values.                                                                            |
| `Export-Exporter`           | Shows the final column structure expected for the CSV upload file.                                                               |

Although the spreadsheet template is useful for learning the format and for data owners to enter their data, our goal in a reproducible data rescue workflow is to avoid manually copying cleaned data into Excel or Google Sheets (and to avoid proprietary software in general!).
Manual copying is difficult to reproduce, time-intensive, and can introduce new errors at the final step.

Instead, in this chapter we will use R to prepare and check a near-final DataStream-formatted table.
To do this, **we will use DataStream's machine-readable schema, which is provided in JSON format (JavaScript Object Notation).**
JSON is a text-based format for storing and exchanging structured data.
This JSON schema contains many of the same rules represented in the spreadsheet template, including field names, field types, allowed values, required fields, and conditional field rules.
*You do not need to understand how to read or write JSON as part of this chapter, but should know that it is a common format for dictating the structure of a data schema* (YAML and XML are other common, human-readable alternatives that you may have encountered).

We will use R to read the JSON schema, extract the parts that are useful for validation, and check whether our formatted dataset follows DataStream's rules before exporting a final CSV.

::: callout-note
### Pre-upload checks are not the final DataStream validator

The checks in this chapter are designed to catch common problems before upload.
**They do not replace DataStream's own upload validation, which occurs on the backend of their platform when data owners upload files.**

You can think of our validation checks in this chapter as a way to identify issues earlier in your reproducible R workflow, and so that data owners do not encounter upload issues related to schema violations.
:::

<br>

## R Workflow for Validating A Dataset Against the DataStream Data Schema

The following R workflow serves as an interactive and reproducible way for you to check if a dataset adheres to the DataStream data schema.
You can incorporate and adapt the code from this section to help validate a real dataset against the DataStream data schema.

**The code in this chapter is meant to be run in your own RStudio session or project environment. The online guidebook shows the code and explains the workflow, but it does not provide a live R session. To follow along, you can copy the code chunks into RStudio or download the .qmd file for this chapter (linked here) and run the code chunks from there.**

The workflow encompasses a portion of the steps in the diagram below.
*Note that a full data rescue workflow would begin with preserving raw files and cleaning or reshaping the original data.* In this chapter, we begin later in the workflow and assume that the data have already been cleaned and reshaped into a table that is close to DataStream format.
Our goal in this chapter is to check that near-final table against the DataStream schema, run final QA/QC (Quality Assurance and Quality Control) checks, and export a final CSV that can be shared with the data owner for upload to the DataStream platform.

```{mermaid}
%%| fig-cap: "Workflow for preparing a near-DataStream-formatted dataset for data owner review. This chapter "
flowchart TD
  A[(Receive raw data files)] --> B[Preserve raw files]
  B --> C[Clean, organize,<br/>and perform initial QA/QC]
  C --> D[Format into DataStream schema]
  D --> E{Pass schema validation?}
  E -- No --> F[Revisit cleaning workflow]
  F --> C
  E -- Yes --> G{Pass final QA/QC checks?}
  G -- No --> F
  G -- Yes --> H[Export final CSV]
  H --> I[Share final CSV<br/>with data owner]
  I --> J[(Data owner uploads<br/>to DataStream)]

  classDef prior fill:#f2f2f2,stroke:#999,color:#666;
  classDef current fill:#ffffff,stroke:#333,stroke-width:2px;
  class A,B,C prior;
  class D,E,F,G,H,I current;
```

### Importing the DataStream Data Schema into R

Because the data schema is in JSON format, we will use the `jsonlite` package to read the DataStream data schema into R, as well as other packages for manipulating data.
The following code will load each of the required R packages and install them for you if they are not installed already.

```{r}
packages <- c(
  "jsonlite", # Reads JSON files, including the DataStream schema
  "dplyr",  # Provides data manipulation functions such as filter() and mutate()
  "purrr",  # Provides mapping functions used to iterate over schema fields
  "stringr", # Provides string helpers for detecting blanks and checking patterns
  "readr", # Writes CSV files
  "tibble" # Creates tidy data frames called tibbles
)

missing_packages <- packages[
  !vapply(packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  install.packages(
    missing_packages,
    repos = "https://cloud.r-project.org"
  )
}

invisible(
  lapply(packages, library, character.only = TRUE)
)
```

Next, we can create folders for reference files and outputs.

```{r}
# Create a folder for reference files, such as the downloaded DataStream schema, and a folder for outputs
# showWarnings = FALSE prevents R from warning us if the folder already exists
dir.create("data_reference", showWarnings = FALSE)
dir.create("outputs", showWarnings = FALSE)
```

Now let's download the machine-readable JSON version of the current DataStream schema.
The schema is accessed directly from the DataStream website, ensuring that we are accessing the most up-to-date version.

```{r}
# Download the current DataStream schema from DataStream's public schema URL
download.file(
  url = "https://datastream.org/schema",
  destfile = "data_reference/datastream_schema.json", # Save it to your new data_reference folder
  mode = "wb"
)
```

Read the schema into R with the `jsonlite` package.
The schema is stored as a text-based JSON file, and `jsonlite::read_json()` converts it into a nested list that we can inspect and use for validation checks.

```{r}
# Read the downloaded JSON schema into R
# simplifyVector = FALSE preserves the nested structure of the schema
# This makes it easier to extract schema rules later
datastream_schema <- jsonlite::read_json(
  path = "data_reference/datastream_schema.json", 
  simplifyVector = FALSE
)
```

You should now see the `datastream_schema` object stored in your R environment as a list.
The schema also includes a version number, and it is a good idea to record this version in your project notes so that others know which version of the schema was used during validation.

```{r}
# Extract the schema version from the "$comment" field
# This allows us to document which schema version was used
schema_version <- datastream_schema[["$comment"]]

# Print the schema version
schema_version
```

<br>

### Extracting Schema Information

The code below extracts the parts of the schema we need for pre-upload checks.
You do not need to understand the full JSON structure of the schema, but should understand the validation logic being built from it.
The goal is to pull out schema information that helps us determine whether our dataset follows DataStream's formatting rules.

```{r}
# Define a helper operator
# If x is NULL, return y. Otherwise, return x
# This helps us avoid errors when a schema element is missing
`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

# Extract the main part of the schema that contains field definitions
# In this schema, the core object is stored inside allOf[[1]]
schema_core <- datastream_schema$allOf[[1]]

# Extract the schema properties for each field
# Each property corresponds to a DataStream field, such as DatasetName,
# MonitoringLocationID, ActivityStartDate, or CharacteristicName
schema_properties <- schema_core$properties

# Extract the names of all fields defined in the DataStream schema
schema_fields <- names(schema_properties)


# Extract the fields that are required by the schema
# If the schema does not include a required section, this returns an empty character vector
required_fields <- schema_core$required %||% character(0)

# Extract dependency rules for the schema
# These are one type of conditional rule:
# 'Dependencies' are cases where if one field is populated, then one
# or more related fields must also be populated
schema_dependencies <- schema_core$dependencies %||% list()
```

After running the above, you should see each of the new objects in your R environment as lists that have now been extracted from the downloaded schema.

Inspect each of the new objects.
For example, the object `schema_fields` contains the names of all fields defined by DataStream.

```{r}
schema_fields
```

The object `required_fields` contains fields that are required by the schema.

```{r}
required_fields
```

<br>

### Create a Field Summary

Now we will create a summary table that describes each DataStream field.
If you are interested in fully understanding what this function is doing, in-line comments are included throughout.

```{r}
# Helper function for extracting text values from the schema
# Some schema fields have a single text value, some have multiple values,
# and some do not have the value at all
get_chr <- function(x) {
  if (is.null(x)) {
    # If the schema field does not contain this information,
    # return NA so the summary table still has a value
    NA_character_
  } else {
    # If the schema field contains one or more values,
    # collapse them into a single character string
    paste(unlist(x), collapse = ", ")
  }
}

# Helper function for extracting numeric values from the schema
# This is used for fields such as minimum, maximum, minLength, and maxLength
get_num <- function(x) {
  if (is.null(x)) {
    # If the schema field does not include a numeric limit,
    # return NA.
    NA_real_
  } else {
    # Extract the first numeric value.
    # For this schema, limits such as minimum and maximum
    # are expected to contain a single value.
    as.numeric(x[[1]])
  }
}

# Create a summary table of the DataStream schema fields.
# Each element of schema_properties corresponds to one DataStream field,
# such as DatasetName, MonitoringLocationID, ActivityStartDate,
# CharacteristicName, or ResultUnit.
schema_field_summary <- purrr::imap_dfr(
  schema_properties,
  function(field_schema, field_name) {
    
    # For each DataStream field, extract the pieces of information
    # that are useful for validation.
    tibble::tibble(
      # Name of the DataStream field
      field = field_name,
      
      # Expected data type, such as string or number
      type = get_chr(field_schema$type),
      
      # Expected format, such as date
      format = get_chr(field_schema$format),
      
      # Minimum numeric value, if defined
      minimum = get_num(field_schema$minimum),
      
      # Maximum numeric value, if defined
      maximum = get_num(field_schema$maximum),
      
      # Minimum number of characters, if defined
      min_length = get_num(field_schema$minLength),
      
      # Maximum number of characters, if defined
      max_length = get_num(field_schema$maxLength),
      
      # TRUE if the field has a controlled vocabulary
      # stored as "enum" value in the schema
      has_allowed_values = !is.null(field_schema$enum)
    )
  }
)

# View the resulting field summary table
schema_field_summary
```

This summary table gives us a compact view of the schema fields, including field type, date format, numeric limits, and whether a field has a controlled vocabulary.

<br>

### Extract Allowed Values

Many DataStream fields use controlled vocabularies.
These are represented in the schema as allowed values.
For example, `MonitoringLocationType`, `ActivityType`, `ActivityMediaName`, `CharacteristicName`, and `ResultUnit` all use controlled vocabularies.
The code below will extract all controlled vocabulary values from the schema.

```{r}
# Extract all controlled vocabulary values from the schema
# In JSON schema, controlled vocabularies are stored as "enum" values
allowed_values <- purrr::imap_dfr(
  schema_properties,
  function(field_schema, field_name) {
    
    # If a field does not have an enum, it does not have allowed values
    # that we need to extract here
    if (is.null(field_schema$enum)) {
      return(tibble::tibble())
    }
    
    # If the field does have an enum, create a table with:
    #   1. the field name; and
    #   2. each allowed value for that field
    tibble::tibble(
      field = field_name,
      allowed_value = as.character(unlist(field_schema$enum))
    )
  }
)

# View the table of allowed values.
allowed_values
```

In the `allowed_values` table, notice that certain fields will only accept one or a few values (e.g., `MonitoringLocationHorizontalCoordinateReferenceSystem` only accepts `NAD27`,`NAD83`,`UNKWN`, `WGS72`, or `WGS84`), while other fields may accept many values (e.g., `CharacteristicName` has \>1500 allowed values).

```{r}
# Filter the allowed values table to count the number of allowed values for CharacteristicName
allowed_values |>
  dplyr::filter(field == "CharacteristicName") |>
  nrow()
```

<br>

### Example Dataset

Suppose we have reshaped a small water quality dataset as we prepare it for DataStream.
The table below is close to DataStream format, but intentionally includes errors.
Let's use this example dataset to show how validation checks would work on an example dataset.

```{r}
# Create a small example dataset that resembles a DataStream-formatted export.
# This dataset intentionally includes several common issues seen during
# data standardization and upload preparation.

datastream_export <- tibble::tribble(
  ~DatasetName, ~MonitoringLocationID, ~MonitoringLocationName,
  ~MonitoringLocationLatitude, ~MonitoringLocationLongitude,
  ~MonitoringLocationHorizontalCoordinateReferenceSystem,
  ~MonitoringLocationType, ~ActivityType, ~ActivityMediaName,
  ~ActivityStartDate, ~ActivityStartTime, ~ActivityStartTimeZone,
  ~CharacteristicName, ~ResultValue, ~ResultUnit, ~ResultValueType,
  ~ResultDetectionCondition,
  ~ResultDetectionQuantitationLimitMeasure,
  ~ResultDetectionQuantitationLimitUnit,
  ~ResultDetectionQuantitationLimitType,
  ~SourceFile, ~SourceSheet,
  
  # Row 1: A valid-looking temperature observation.
  # This row gives us one example that should mostly pass the basic checks.
  "Example Mountain Lake Monitoring", "ML-01", "Clear Lake",
  49.850, -123.100, "WGS84",
  "Lake/Pond", "Field Msr/Obs", "Surface Water",
  "2024-07-15", "10:30:00", "-07:00",
  "Temperature, water", 14.2, "deg C", "Actual",
  NA_character_,
  NA_real_, NA_character_, NA_character_,
  "lake_data_2024.xlsx", "July field data",
  
  # Row 2: Same apparent location, but with slightly different coordinates.
  # This can happen when coordinates vary across files or years because of
  # rounding, GPS precision, or inconsistent transcription.
  # This row also includes an invalid date format, invalid time format,
  # and a CharacteristicName that may not exactly match DataStream's vocabulary.
  "Example Mountain Lake Monitoring", "ML-01", "Clear Lake",
  49.85012, -123.10009, "WGS84",
  "Lake/Pond", "Field Msr/Obs", "Surface Water",
  "07/15/2024", "10:35 AM", "-07:00",
  "Dissolved oxygen", 8.9, "mg/L", "Actual",
  NA_character_,
  NA_real_, NA_character_, NA_character_,
  "lake_data_2024.xlsx", "July field data",
  
  # Row 3: Same apparent location again, but with different capitalization
  # in both the location ID and name.
  # Because location IDs and names are case sensitive, this could create
  # duplicate locations during upload if not corrected.
  # This row also includes a negative pH value, which is scientifically invalid
  # even if it may not be caught by basic schema checks.
  "Example Mountain Lake Monitoring", "ml-01", "Clear lake",
  49.850, -123.100, "WGS84",
  "Lake/Pond", "Field Msr/Obs", "Surface Water",
  "2024-07-15", "10:40:00", "-07:00",
  "pH", -1, "None", "Actual",
  NA_character_,
  NA_real_, NA_character_, NA_character_,
  "lake_data_2024.xlsx", "July field data",
  
  # Row 4: Missing coordinates and missing time zone.
  # Missing coordinates and time zones are common issues in raw data.
  # This row also uses "Water" instead of the DataStream-controlled value
  # "Surface Water".
  "Example Mountain Lake Monitoring", "ML-02", "Mirror Lake",
  NA_real_, NA_real_, "WGS84",
  "Lake/Pond", "Sample-Routine", "Water",
  "2024-07-16", "11:20:00", NA_character_,
  "Specific conductance", 125, "uS/cm", "Actual",
  NA_character_,
  NA_real_, NA_character_, NA_character_,
  "lake_data_2024.xlsx", "July chemistry",
  
  # Row 5: A censored result.
  # The result is flagged as below a detection/quantification limit,
  # but the limit value, limit unit, and limit type are missing.
  # This is a common detection-limit problem in historical datasets.
  "Example Mountain Lake Monitoring", "ML-02", "Mirror Lake",
  49.910, -123.050, "WGS84",
  "Lake/Pond", "Sample-Routine", "Surface Water",
  "2024-07-16", "11:25:00", "-07:00",
  "Ammonia and ammonium", NA_real_, NA_character_, "Actual",
  "Below Detection/Quantification Limit",
  NA_real_, NA_character_, NA_character_,
  "lab_results_2024.xlsx", "Nutrients",
  
  # Row 6: A negative hardness value.
  # This kind of issue is not just a formatting problem; it requires
  # scientific QA/QC because hardness should not be less than or equal to zero.
  "Example Mountain Lake Monitoring", "ML-03", "Long Lake",
  49.960, -123.020, "WGS84",
  "Lake/Pond", "Sample-Routine", "Surface Water",
  "2024-07-17", "09:15:00", "-07:00",
  "Hardness, Ca, Mg", -5, "mg/L", "Actual",
  NA_character_,
  NA_real_, NA_character_, NA_character_,
  "lab_results_2024.xlsx", "General chemistry"
)
```

This example includes several possible issues: - Dissolved oxygen may not exactly match the allowed DataStream value.
- One date uses MM/DD/YYYY instead of YYYY-MM-DD.
- One row has a detection condition but no result value or detection limit information.
- Some sample records may require additional analytical method information before upload.

<br>

### Validation Helper Functions

Before checking how our example dataset adheres to the schema, we will create a few "helper functions." The purpose of these functions is to make validation easier to read and interpret.

```{r}
# Helper function for identifying blank values
# This treats both NA values and empty strings as blank
is_blank <- function(x) {
  is.na(x) | stringr::str_trim(as.character(x)) == ""
}

# Helper function for creating a validation issue table
# Each validation check will use this function to return issues
# in the same format
make_issue <- function(data, rows, field, issue) {
  
  # If there are no problem rows, return an empty tibble
  if (length(rows) == 0) {
    return(tibble::tibble())
  }
  
  # If the field exists in the dataset, extract the values from the problem rows
  # If the field does not exist, return NA values instead
  values <- if (field %in% names(data)) {
    as.character(data[[field]][rows])
  } else {
    rep(NA_character_, length(rows))
  }
  
  # Return a standardized validation report table
  tibble::tibble(
    row = rows,
    field = field,
    value = values,
    issue = issue
  )
}
```

Now that we've defined these helper functions, each validation check will return a table with: - The row number - The field containing the issue - The value currently in that field - A short description of the issue

<br>

### Check 1: Unexpected Columns

```{r}
# Check whether the dataset contains columns that are not defined
# in the DataStream schema
check_extra_columns <- function(data, schema_fields) {
  
  # Identify columns in the dataset that are not in the schema
  extra_cols <- setdiff(names(data), schema_fields)
  
  # If there are no extra columns, return an empty tibble
  if (length(extra_cols) == 0) {
    return(tibble::tibble())
  }
  
  # If extra columns exist, report them as validation issues
  tibble::tibble(
    row = NA_integer_,
    field = extra_cols,
    value = NA_character_,
    issue = "Column is not defined in the DataStream schema."
  )
}

# Run the extra-column check on the example dataset
check_extra_columns(datastream_export, schema_fields)
```

No extra columns exist in this example dataset.

### Check 2: Required Fields

Now let's check that all required fields are populated.

```{r}
# Check whether required DataStream fields are present and populated.
check_required_fields <- function(data, required_fields) {
  
  # Identify required fields that are missing entirely from the dataset.
  missing_required <- setdiff(required_fields, names(data))
  
  # Only create missing-column issues if there are actually missing columns.
  missing_issues <- if (length(missing_required) > 0) {
    
    tibble::tibble(
      row = NA_integer_,
      field = missing_required,
      value = NA_character_,
      issue = "Required field is missing from the dataset."
    )
    
  } else {
    
    # If no required columns are missing, return an empty tibble.
    tibble::tibble()
  }
  
  # Identify required fields that are present in the dataset.
  present_required <- intersect(required_fields, names(data))
  
  # For each required field that is present, check whether any rows are blank.
  blank_issues <- purrr::map_dfr(present_required, function(field) {
    
    # Find rows where this required field is blank.
    bad_rows <- which(is_blank(data[[field]]))
    
    # Create a validation issue for each blank required value.
    make_issue(
      data = data,
      rows = bad_rows,
      field = field,
      issue = "Required field is blank."
    )
  })
  
  # Combine missing-column issues and blank-value issues.
  dplyr::bind_rows(missing_issues, blank_issues)
}

# Run the required-field check on the example dataset.
check_required_fields(datastream_export, required_fields)
```

All required fields are populated.

<br>

### Check 3: Allowed Values

Many DataStream fields use controlled vocabularies.
For these fields, any non-blank value should match one of the allowed values in the schema.
This check is especially useful for fields like: - MonitoringLocationType - ActivityType - ActivityMediaName - CharacteristicName - ResultUnit - ResultValueType - ResultDetectionCondition

```{r}
# Check whether controlled vocabulary fields use allowed DataStream values.
check_allowed_values <- function(data, allowed_values) {
  
  # Identify fields that both:
  #   1. have allowed values in the schema; and
  #   2. are present in the dataset.
  fields_to_check <- intersect(
    unique(allowed_values$field),
    names(data)
  )
  
  # Check each controlled vocabulary field one at a time.
  purrr::map_dfr(fields_to_check, function(field) {
    
    # Pull the allowed values for the current field.
    allowed <- allowed_values |>
      dplyr::filter(.data$field == .env$field) |>
      dplyr::pull(allowed_value)
    
    # Find rows where the value is not blank
    # but also not included in the allowed values.
    bad_rows <- which(
      !is_blank(data[[field]]) &
        !(as.character(data[[field]]) %in% allowed)
    )
    
    # Create validation issues for invalid values.
    make_issue(
      data = data,
      rows = bad_rows,
      field = field,
      issue = "Value is not listed as an allowed value in the DataStream schema."
    )
  })
}

# Run the allowed-value check on the example dataset.
check_allowed_values(datastream_export, allowed_values)
```

It appears that we have a CharacteristicName that does not match any of the allowed values in the schema.

### Check 4: Numeric Fields and Numeric Ranges

Some fields must be numeric.
The schema also defines minimum and maximum values for some numeric fields.

For example, latitude must be between 0 and 90, and longitude must be between -180 and 180.

```{r}
# Check whether numeric fields contain numeric values
# and whether those values fall within schema-defined ranges
check_numeric_ranges <- function(data, schema_field_summary) {
  
  # Identify numeric fields that are present in the dataset
  numeric_fields <- schema_field_summary |>
    dplyr::filter(type %in% c("number", "integer")) |>
    dplyr::filter(field %in% names(data))
  
  # Check each numeric field one at a time
  purrr::map_dfr(seq_len(nrow(numeric_fields)), function(i) {
    
    # Extract the field name and its allowed numeric range
    field <- numeric_fields$field[i]
    min_value <- numeric_fields$minimum[i]
    max_value <- numeric_fields$maximum[i]
    
    # Try to convert the field values to numeric
    # suppressWarnings() avoids printing warnings for non-numeric values
    x <- suppressWarnings(as.numeric(data[[field]]))
    
    # Find rows where:
    #   1. the value is not blank; and
    #   2. the value cannot be converted to numeric, or
    #   3. the value falls below the minimum, or
    #   4. the value exceeds the maximum
    bad_rows <- which(
      !is_blank(data[[field]]) &
        (
          is.na(x) |
            (!is.na(min_value) & x < min_value) |
            (!is.na(max_value) & x > max_value)
        )
    )
    
    # Create validation issues for invalid numeric values
    make_issue(
      data = data,
      rows = bad_rows,
      field = field,
      issue = "Value is not numeric or falls outside the range defined in the DataStream schema."
    )
  })
}

# Run the numeric range check on the example dataset
check_numeric_ranges(datastream_export, schema_field_summary)
```

### Check 5: Date, Time, and Time Zone Formats

DataStream expects dates, times, and time zones to use specific formats.
A common data rescue problem is that dates may be stored inconsistently across source files.

```{r}
# Check whether date, time, and time zone fields use expected formats
check_date_time_formats <- function(data, schema_field_summary) {
  
  # Create a list to store validation issues from date, time, and time zone checks
  issues <- list()
  
  # Identify fields that the schema marks as date fields
  date_fields <- schema_field_summary |>
    dplyr::filter(format == "date") |>
    dplyr::pull(field)
  
  # Keep only date fields that are present in the dataset
  date_fields <- intersect(date_fields, names(data))
  
  # Check whether date fields use YYYY-MM-DD format.
  issues[["dates"]] <- purrr::map_dfr(date_fields, function(field) {
    
    # Find rows where the date is not blank and does not match YYYY-MM-DD
    bad_rows <- which(
      !is_blank(data[[field]]) &
        !stringr::str_detect(as.character(data[[field]]), "^\\d{4}-\\d{2}-\\d{2}$")
    )
    
    # Create validation issues for invalid dates
    make_issue(
      data = data,
      rows = bad_rows,
      field = field,
      issue = "Date should use YYYY-MM-DD format."
    )
  })
  
  # Define the time fields we want to check, if they are present
  time_fields <- intersect(
    c("ActivityStartTime", "ActivityEndTime", "AnalysisStartTime"),
    names(data)
  )
  
  # Regular expression for valid time values.
  # This accepts HH:MM, HH:MM:SS, or HH:MM:SS.sss.
  time_pattern <- "^(2[0-3]|[01][0-9]):[0-5][0-9](:[0-5][0-9](\\.[0-9]{1,3})?)?$"
  
  # Check whether time fields match the expected time pattern
  issues[["times"]] <- purrr::map_dfr(time_fields, function(field) {
    
    # Find rows where the time is not blank and does not match the pattern
    bad_rows <- which(
      !is_blank(data[[field]]) &
        !stringr::str_detect(as.character(data[[field]]), time_pattern)
    )
    
    # Create validation issues for invalid times
    make_issue(
      data = data,
      rows = bad_rows,
      field = field,
      issue = "Time should use HH:MM, HH:MM:SS, or HH:MM:SS.sss format."
    )
  })
  
  # Define the time zone fields we want to check, if they are present
  timezone_fields <- intersect(
    c("ActivityStartTimeZone", "ActivityEndTimeZone", "AnalysisStartTimeZone"),
    names(data)
  )
  
  # Regular expression for valid time zone values
  # This accepts Z or offsets such as -07:00 or +05:30
  timezone_pattern <- "^(Z|[+-](0[0-9]|1[0-9]|2[0-4]):[0-5][0-9])$"
  
  # Check whether time zone fields match the expected pattern.
  issues[["timezones"]] <- purrr::map_dfr(timezone_fields, function(field) {
    
    # Find rows where the time zone is not blank and does not match the pattern
    bad_rows <- which(
      !is_blank(data[[field]]) &
        !stringr::str_detect(as.character(data[[field]]), timezone_pattern)
    )
    
    # Create validation issues for invalid time zones
    make_issue(
      data = data,
      rows = bad_rows,
      field = field,
      issue = "Time zone should use Z or +/-HH:MM format."
    )
  })
  
  # Combine all date, time, and time zone issues into one table
  dplyr::bind_rows(issues)
}

# Run the date, time, and time zone checks on the example dataset
check_date_time_formats(datastream_export, schema_field_summary)
```

<br>

### Check 6: Field Dependencies

Some fields become required when another field is populated.
These are called dependencies.

For example, if `ResultValue` is populated, then `ResultUnit` is also required.
If `ActivityDepthHeightMeasure` is populated, then `ActivityDepthHeightUnit` is also required.

We can extract these dependencies from the schema.

```{r}
# Convert schema dependencies into a table that is easier to inspect
# Each row says: when trigger_field is populated,
# dependent_field should also be populated
dependency_table <- purrr::imap_dfr(
  schema_dependencies,
  function(dependent_fields, trigger_field) {
    
    tibble::tibble(
      trigger_field = trigger_field,
      dependent_field = as.character(unlist(dependent_fields))
    )
  }
)

# View the dependency table
dependency_table
```

Now we can check whether dependent fields are populated when their trigger fields are populated.

```{r}
# Check whether dependent fields are populated when their trigger fields are populated
check_dependencies <- function(data, dependency_table) {
  
  # If there are no dependencies, return an empty tibble
  if (nrow(dependency_table) == 0) {
    return(tibble::tibble())
  }
  
  # Check each dependency rule one at a time
  purrr::map_dfr(seq_len(nrow(dependency_table)), function(i) {
    
    # Extract the trigger field and dependent field for this rule
    trigger_field <- dependency_table$trigger_field[i]
    dependent_field <- dependency_table$dependent_field[i]
    
    # If the trigger field is not in the dataset, skip this rule
    if (!trigger_field %in% names(data)) {
      return(tibble::tibble())
    }
    
    # Identify rows where the trigger field is populated
    trigger_rows <- which(!is_blank(data[[trigger_field]]))
    
    # If the trigger field is never populated, there is nothing to check
    if (length(trigger_rows) == 0) {
      return(tibble::tibble())
    }
    
    # If the dependent field is missing entirely, report it as an issue
    # for each row where the trigger field is populated
    if (!dependent_field %in% names(data)) {
      return(
        tibble::tibble(
          row = trigger_rows,
          field = dependent_field,
          value = NA_character_,
          issue = paste0(
            "Field is required when ",
            trigger_field,
            " is populated, but this column is missing."
          )
        )
      )
    }
    
    # Among rows where the trigger field is populated,
    # identify rows where the dependent field is blank
    bad_rows <- trigger_rows[
      is_blank(data[[dependent_field]][trigger_rows])
    ]
    
    # Create validation issues for missing dependent values
    make_issue(
      data = data,
      rows = bad_rows,
      field = dependent_field,
      issue = paste0(
        "Required when ",
        trigger_field,
        " is populated."
      )
    )
  })
}

# Run the dependency check on the example dataset
check_dependencies(datastream_export, dependency_table)
```

<br>

### Check 7: Result Values and Detection Conditions

A common DataStream issue involves the relationship between `ResultValue` and `ResultDetectionCondition`.
For a single observation, you should generally provide either:

-   A measured value in `ResultValue`, or
-   A detection condition in `ResultDetectionCondition`.

You should not populate both for the same observation, and you should not leave both blank.
`ResultDetectionCondition` is a required field when `ResultValue` is blank.

```{r}
# Check the relationship between ResultValue and ResultDetectionCondition
check_result_value_detection_condition <- function(data) {
  
  # These are the two fields needed for this check
  required_fields <- c("ResultValue", "ResultDetectionCondition")
  
  # If either field is missing, skip this check
  if (!all(required_fields %in% names(data))) {
    return(tibble::tibble())
  }
  
  # Identify rows where both ResultValue and ResultDetectionCondition are blank
  both_blank <- which(
    is_blank(data$ResultValue) &
      is_blank(data$ResultDetectionCondition)
  )
  
  # Identify rows where both ResultValue and ResultDetectionCondition are populated
  both_populated <- which(
    !is_blank(data$ResultValue) &
      !is_blank(data$ResultDetectionCondition)
  )
  
  # Combine both types of issues into one table
  dplyr::bind_rows(
    make_issue(
      data = data,
      rows = both_blank,
      field = "ResultValue",
      issue = "Either ResultValue or ResultDetectionCondition should be populated."
    ),
    make_issue(
      data = data,
      rows = both_populated,
      field = "ResultDetectionCondition",
      issue = "Do not populate ResultDetectionCondition when ResultValue is also populated."
    )
  )
}

# Run the ResultValue / ResultDetectionCondition check
check_result_value_detection_condition(datastream_export)
```

<br>

### Create a Validation Report

Now combine the checks into a single validation function.

```{r}
# Run all pre-upload validation checks and combine the results
validate_datastream_preupload <- function(data,
                                          schema_fields,
                                          required_fields,
                                          allowed_values,
                                          schema_field_summary,
                                          dependency_table) {
  
  dplyr::bind_rows(
    # Check whether all columns are recognized by the DataStream schema
    check_extra_columns(data, schema_fields),
    
    # Check whether required fields are present and populated
    check_required_fields(data, required_fields),
    
    # Check whether controlled vocabulary fields use allowed values
    check_allowed_values(data, allowed_values),
    
    # Check numeric fields and schema-defined numeric ranges
    check_numeric_ranges(data, schema_field_summary),
    
    # Check date, time, and time zone formatting
    check_date_time_formats(data, schema_field_summary),
    
    # Check schema-defined field dependencies
    check_dependencies(data, dependency_table),
    
    # Check the relationship between result values and detection conditions
    check_result_value_detection_condition(data)
  ) |>
    # Sort the validation report by row and field
    dplyr::arrange(row, field)
}
```

And run the validation checks.

```{r}
# Run the combined validation function on the example dataset
validation_report <- validate_datastream_preupload(
  data = datastream_export,
  schema_fields = schema_fields,
  required_fields = required_fields,
  allowed_values = allowed_values,
  schema_field_summary = schema_field_summary,
  dependency_table = dependency_table
)

# View the validation report
validation_report
```

```{r}
# If the validation report contains one or more rows,
# write the report to a CSV file
if (nrow(validation_report) > 0) {
  
  # Save the validation report to the outputs folder so it can be reviewed
  readr::write_csv(
    validation_report,
    "outputs/datastream_validation_report.csv"
  )
  
  # In a real data processing script, you may want to stop the pipeline here
  message(
    "DataStream pre-upload checks found issues. ",
    "Review outputs/datastream_validation_report.csv before exporting the final CSV."
  )
  
  # In a real pipeline, you could use stop() instead of message():
  # stop("DataStream pre-upload checks failed. Review outputs/datastream_validation_report.csv.")
}
```

If the validation report is populated with issues, you will receive an error here that points to the saved validation report in the outputs folder.

<br>

### Review: What These Checks Do and Do Not Do

These checks help determine whether the dataset is likely to conform to the DataStream schema.
They are useful for catching common issues before upload.

| Question                                                    | Addressed by these checks? |
|-----------------------------------------------|-------------------------|
| Are unexpected columns present?                             | Yes                        |
| Are required fields populated?                              | Yes                        |
| Are controlled vocabulary values valid?                     | Yes                        |
| Are numeric values within schema-defined ranges?            | Yes                        |
| Are dates, times, and time zones formatted consistently?    | Yes                        |
| Are common dependent fields populated when needed?          | Yes                        |
| Does this fully reproduce DataStream's upload validator?    | No                         |
| Did the cleaning script preserve all original observations? | No                         |
| Were duplicate observations introduced during joins?        | No                         |
| Are values scientifically reasonable?                       | Only partly                |

Final validation happens during the DataStream upload process.
In addition, schema checks should be paired with project-specific QA/QC checks that confirm the cleaned data are scientifically reasonable and correctly derived from the raw files (see linked chapter).

## Final QA/QC Checks After Schema Validation

Before exporting the final CSV, we should run a final set of QA/QC checks.
**Importantly, these checks are not meant to replace the data cleaning and QA/QC work that should already have occurred earlier in the data rescue workflow.** Instead, they provide a final review of common issues that may remain after the dataset has been cleaned, reshaped, and mapped to DataStream fields.

At this stage, we assume that the dataset has already been substantially cleaned and organized.
The goal is to check the formatted table for issues that may not be caught by schema validation alone, such as near-duplicate locations, inconsistent coordinates, missing time zones, incomplete detection-limit information, or scientifically implausible values.

If these checks identify problems, the preferred approach is to fix them upstream in the R cleaning script, then rerun the schema validation and final QA/QC checks.

As covered in the data cleaning chapter, DataStream often see issues like:

-   Duplicate or near-duplicate monitoring locations
-   Different coordinates associated with the same location
-   Missing coordinates or time zones
-   Implausible or impossible values
-   Incomplete detection-limit information
-   Inconsistent naming conventions across files or sheets

These issues are not always schema validation errors.
Instead, they are **QA/QC issues** that should be reviewed before the final CSV is shared with the data owner.

<br>

### Check for Location Issues

Duplicate locations can be created when the same site is represented with different spelling, capitalization, or coordinates.
Because DataStream treats location IDs and names as case sensitive, small differences can matter.

```{r}
# Check for common monitoring location issues
check_location_qaqc <- function(data) {
  
  # If the required location fields are not present, skip this check
  location_fields <- c(
    "MonitoringLocationID",
    "MonitoringLocationName",
    "MonitoringLocationLatitude",
    "MonitoringLocationLongitude"
  )
  
  if (!all(location_fields %in% names(data))) {
    return(tibble::tibble())
  }
  
  # Check for location IDs that differ only by capitalization
  # For example, "ML-01" and "ml-01" may refer to the same location
  # but would be treated as different values
  id_case_issues <- data |>
    dplyr::mutate(
      row = dplyr::row_number(),
      id_lower = stringr::str_to_lower(MonitoringLocationID)
    ) |>
    dplyr::group_by(id_lower) |>
    dplyr::filter(dplyr::n_distinct(MonitoringLocationID) > 1) |>
    dplyr::ungroup() |>
    dplyr::transmute(
      row = row,
      field = "MonitoringLocationID",
      value = as.character(MonitoringLocationID),
      issue = "Location IDs differ only by capitalization. Review for possible duplicate locations."
    )
  
  # Check for location names that differ only by capitalization
  # For example, "Clear Lake" and "Clear lake" may refer to the same location
  name_case_issues <- data |>
    dplyr::mutate(
      row = dplyr::row_number(),
      name_lower = stringr::str_to_lower(MonitoringLocationName)
    ) |>
    dplyr::group_by(name_lower) |>
    dplyr::filter(dplyr::n_distinct(MonitoringLocationName) > 1) |>
    dplyr::ungroup() |>
    dplyr::transmute(
      row = row,
      field = "MonitoringLocationName",
      value = as.character(MonitoringLocationName),
      issue = "Location names differ only by capitalization. Review for possible duplicate locations."
    )
  
  # Check for cases where the same location ID has more than one
  # latitude or longitude value
  # This can happen because of rounding, changing GPS precision,
  # or inconsistent coordinates across years or files
  coordinate_issues <- data |>
    dplyr::mutate(row = dplyr::row_number()) |>
    dplyr::group_by(MonitoringLocationID) |>
    dplyr::filter(
      dplyr::n_distinct(MonitoringLocationLatitude, na.rm = TRUE) > 1 |
        dplyr::n_distinct(MonitoringLocationLongitude, na.rm = TRUE) > 1
    ) |>
    dplyr::ungroup() |>
    dplyr::transmute(
      row = row,
      field = "MonitoringLocationLatitude/Longitude",
      value = paste(
        MonitoringLocationLatitude,
        MonitoringLocationLongitude,
        sep = ", "
      ),
      issue = "The same MonitoringLocationID has more than one coordinate pair. Review for rounding, precision, or transcription issues."
    )
  
  # Check for missing coordinates
  missing_coordinate_issues <- data |>
    dplyr::mutate(row = dplyr::row_number()) |>
    dplyr::filter(
      is_blank(MonitoringLocationLatitude) |
        is_blank(MonitoringLocationLongitude)
    ) |>
    dplyr::transmute(
      row = row,
      field = "MonitoringLocationLatitude/Longitude",
      value = paste(
        MonitoringLocationLatitude,
        MonitoringLocationLongitude,
        sep = ", "
      ),
      issue = "Latitude or longitude is missing. Coordinates should be reviewed before upload."
    )
  
  # Combine all location QA/QC issues into one table
  dplyr::bind_rows(
    id_case_issues,
    name_case_issues,
    coordinate_issues,
    missing_coordinate_issues
  )
}

# Run the location QA/QC check
check_location_qaqc(datastream_export)
```

<br>

### Check for Missing Time Zones

Missing time zones are common in raw data.
They are especially important when time values are included.

```{r}
# Check for missing time zones when an activity start time is provided
check_timezone_qaqc <- function(data) {
  
  # Skip this check if the relevant fields are not present
  timezone_fields <- c("ActivityStartTime", "ActivityStartTimeZone")
  
  if (!all(timezone_fields %in% names(data))) {
    return(tibble::tibble())
  }
  
  # Find rows where a start time is provided but the time zone is blank
  bad_rows <- which(
    !is_blank(data$ActivityStartTime) &
      is_blank(data$ActivityStartTimeZone)
  )
  
  # Return a standardized issue table
  make_issue(
    data = data,
    rows = bad_rows,
    field = "ActivityStartTimeZone",
    issue = "ActivityStartTime is populated, but ActivityStartTimeZone is missing."
  )
}

# Run the time zone QA/QC check
check_timezone_qaqc(datastream_export)
```

<br>

### Check for Impossible or Suspicious Values

Some values may be valid numbers but still scientifically unreasonable.
For example, pH should not be negative, and hardness should not be less than zero.

```{r}
# Check for simple impossible or suspicious values
check_value_qaqc <- function(data) {
  
  # Skip this check if the fields needed for value checks are missing
  value_fields <- c("CharacteristicName", "ResultValue")
  
  if (!all(value_fields %in% names(data))) {
    return(tibble::tibble())
  }
  
  # Convert ResultValue to numeric for checking
  # suppressWarnings() avoids warnings if some values are text
  data_checked <- data |>
    dplyr::mutate(
      row = dplyr::row_number(),
      result_value_num = suppressWarnings(as.numeric(ResultValue))
    )
  
  # Check for pH values outside the usual possible range of 0 to 14
  ph_issues <- data_checked |>
    dplyr::filter(
      CharacteristicName == "pH",
      !is.na(result_value_num),
      result_value_num < 0 | result_value_num > 14
    ) |>
    dplyr::transmute(
      row = row,
      field = "ResultValue",
      value = as.character(ResultValue),
      issue = "pH value is outside the expected range of 0 to 14."
    )
  
  # Check for negative hardness values
  hardness_issues <- data_checked |>
    dplyr::filter(
      stringr::str_detect(CharacteristicName, "Hardness"),
      !is.na(result_value_num),
      result_value_num < 0
    ) |>
    dplyr::transmute(
      row = row,
      field = "ResultValue",
      value = as.character(ResultValue),
      issue = "Hardness value is negative. Review as a likely data quality issue."
    )
  
  # Check for negative values for variables that generally should not be negative
  nonnegative_issues <- data_checked |>
    dplyr::filter(
      CharacteristicName %in% c(
        "Specific conductance",
        "Ammonia and ammonium",
        "Dissolved oxygen",
        "Temperature, water"
      ),
      !is.na(result_value_num),
      result_value_num < 0
    ) |>
    dplyr::transmute(
      row = row,
      field = "ResultValue",
      value = as.character(ResultValue),
      issue = "ResultValue is negative for a characteristic that generally should not be negative. Review before upload."
    )
  
  # Combine all value QA/QC issues into one table
  dplyr::bind_rows(
    ph_issues,
    hardness_issues,
    nonnegative_issues
  )
}

# Run the value QA/QC check
check_value_qaqc(datastream_export)
```

<br>

### Check Detection-Limit Information

Detection-limit issues are common in historical water quality datasets.
If a result is reported as above or below a detection or quantification limit, then the limit value, unit, and type should be reviewed carefully.

```{r}
# Check for incomplete detection-limit information
check_detection_limit_qaqc <- function(data) {
  
  # These fields are needed for this check
  detection_fields <- c(
    "ResultDetectionCondition",
    "ResultDetectionQuantitationLimitMeasure",
    "ResultDetectionQuantitationLimitUnit",
    "ResultDetectionQuantitationLimitType"
  )
  
  # Skip this check if the relevant fields are not present
  if (!all(detection_fields %in% names(data))) {
    return(tibble::tibble())
  }
  
  # Find rows where a detection condition is provided,
  # but one or more detection-limit fields are missing
  incomplete_limit_rows <- which(
    !is_blank(data$ResultDetectionCondition) &
      (
        is_blank(data$ResultDetectionQuantitationLimitMeasure) |
          is_blank(data$ResultDetectionQuantitationLimitUnit) |
          is_blank(data$ResultDetectionQuantitationLimitType)
      )
  )
  
  # Create issues for incomplete detection-limit information
  incomplete_limit_issues <- make_issue(
    data = data,
    rows = incomplete_limit_rows,
    field = "ResultDetectionQuantitationLimitMeasure",
    issue = "ResultDetectionCondition is populated, but detection-limit value, unit, or type is missing."
  )
  
  # Convert the detection-limit measure to numeric
  limit_value <- suppressWarnings(
    as.numeric(data$ResultDetectionQuantitationLimitMeasure)
  )
  
  # Find rows where the detection-limit value is negative
  negative_limit_rows <- which(
    !is_blank(data$ResultDetectionQuantitationLimitMeasure) &
      !is.na(limit_value) &
      limit_value < 0
  )
  
  # Create issues for negative detection-limit values
  negative_limit_issues <- make_issue(
    data = data,
    rows = negative_limit_rows,
    field = "ResultDetectionQuantitationLimitMeasure",
    issue = "Detection-limit value is negative. Review before upload."
  )
  
  # Combine detection-limit QA/QC issues into one table
  dplyr::bind_rows(
    incomplete_limit_issues,
    negative_limit_issues
  )
}

# Run the detection-limit QA/QC check
check_detection_limit_qaqc(datastream_export)
```

<br>

### Create a QA/QC Report

Now, using all of the functions we defined above, combine all of the QA/QC checks into one report that you can save and share with team members.

```{r}
# Run all project-specific QA/QC checks and combine the results
run_project_qaqc <- function(data) {
  
  dplyr::bind_rows(
    # Check for duplicate or inconsistent monitoring locations.
    check_location_qaqc(data),
    
    # Check whether time zones are missing when times are provided
    check_timezone_qaqc(data),
    
    # Check for impossible or suspicious values
    check_value_qaqc(data),
    
    # Check whether detection-limit information is complete
    check_detection_limit_qaqc(data)
  ) |>
    # Sort the QA/QC report by row and field.
    dplyr::arrange(row, field)
}

# Run the QA/QC checks on the example dataset
qaqc_report <- run_project_qaqc(datastream_export)

# View the QA/QC report
qaqc_report
```

If the QA/QC report contains rows, review the issues and return to the original data cleaning workflow.

```{r}
# If the QA/QC report contains one or more rows,
# write the report to a CSV file
if (nrow(qaqc_report) > 0) {
  
  # Save the QA/QC report to the outputs folder so it can be reviewed
  readr::write_csv(
    qaqc_report,
    "outputs/datastream_qaqc_report.csv"
  )
}
```

These QA/QC checks are intentionally simple and assume that the dataset is already nearly formatted to the DataStream schema.
They are meant to flag records that require review, not automatically determine whether a value is correct or incorrect.
In a real data rescue project, the team should discuss the resulting issues with team members, the data owner, and document any assumptions, corrections, or unresolved questions.
