---
title: "Preparing Data for DataStream"
from: markdown+bracketed_spans
editor: 
  markdown: 
    wrap: sentence
---

After preparing metadata for a dataset, we also need to prepare the data itself for publication on DataStream.

In our data rescue projects, the original files we receive will not already follow DataStream's data schema (see the [Data schemas section in the Introduction to Data Chapter](data.qmd#data-schemas) for more information on what a data schema is).
Furthermore, a data owner may provide multiple spreadsheets or CSV files with different column names, inconsistent site names, mixed date formats, non-standard units, censored values, and/or values that require scientific review.

In this chapter, we will continue working with the Decent Mountains National Park example from the previous chapter.
By the end of this chapter, you should be able to:

- Clean the combined table of source data
- Document data cleaning decisions in a change log
- Perform basic scientific validation checks on water quality data
- Format a cleaned water quality dataset into the DataStream data schema
- Run a pre-upload validation check using the [JSON]{.tooltip-term tabindex="0" data-bs-toggle="tooltip" data-bs-title="JSON, or JavaScript Object Notation, is a simple way to write down and organize information using plain text, but that works with almost any computer language. Think of it like a digital label or a neat list that computers use to talk to each other."} version of DataStream's data schema
- Export a CSV and validation report for your team to review

**This chapter is designed primarily as a reference workflow in R.**
You do not need to memorize every function or code block, but can instead focus on understanding the overall sequence and return to the chapter as needed throughout your project.

::: callout-note
### Metadata updates are continuous

In this guide, the workflow is presented in the following sequence: first we prepare **metadata**, then we prepare **data** through cleaning, validation, and formatting.

In a real data rescue project, these steps are iterative.
Cleaning the data often reveals new metadata questions and updates.
When those issues are discovered, they should be added back to the metadata documentation, change log, or list of questions for the data owner.

Metadata should be updated throughout the project as new issues and decisions emerge.
We perform our own update to the example metadata at the end of this chapter.
:::

<br>

## The DataStream Data Schema

DataStream uses a standardized data format based on the WQX (Water Quality Exchange) standard for the exchange of water quality data.
This allows different groups to share results in a common format and increases a dataset's **Interoperability** and **Reusability**, (i.e., the I and R in [FAIR principles](metadata.qmd#sec-FAIR-principles)).

The source data that we receive will almost certainly not already adhere to the DataStream data schema.
In a reproducible data rescue workflow, we want to use R to transform the original files into a table that adheres to the DataStream data schema.
We then export that table as a CSV, which the relevant data owner will upload to DataStream.

### Upload templates and schema validation

DataStream provides spreadsheet upload templates to help users understand how their data should be structured for upload.
These templates include field names, examples, allowed values, and lookup tables.
They are especially helpful for visualizing the expected DataStream format and for understanding how different fields relate to one another.

As you work through this chapter, you can open the current [DataStream upload template](https://datastream.org/en-ca/documentation/data-upload-resources#datastream-upload-templates) (either Excel or Google Sheets) and use it as a visual reference.

The template contains several important sheets to investigate:

| Sheet | Purpose |
|------------------------------------|------------------------------------|
| `Allowed Values-Valeurs` | Lists DataStream fields, whether they are required, optional, or conditional, their field types, examples, and allowed values |
| `CharacteristicName LOOKUP` | Provides additional guidance for `CharacteristicName` values, including fields that may be required for certain characteristics |
| `ResultUnit LOOKUP` | Provides information about valid `ResultUnit` values |
| `Export-Exporter` | Shows the final column structure expected for the CSV upload file |

Although the spreadsheet template is useful for learning the format, our goal in a reproducible workflow is to avoid manually copying cleaned data into Excel or Google Sheets.
Manual copying is difficult to reproduce, time-intensive, and can introduce new errors.

Instead, this chapter will use R to prepare and check a near-final DataStream-formatted table.
At the end of the chapter, we will use DataStream's machine-readable schema, which is provided in JSON format via the DataStream website.
JSON is a text-based format for storing and exchanging structured data, and the JSON schema contains all of the rules represented in the spreadsheet template, including field names, field types, allowed values, required fields, and conditional field rules.

**You do not need to understand how to read or write JSON as part of this chapter.** For this workflow, you only need to know that the JSON schema is a machine-readable reference that we can use in R to check whether a formatted dataset follows DataStream's rules.

::: callout-note
### Pre-upload checks are not the final DataStream validator

The checks in this chapter are designed to catch common problems before upload.
They do not replace DataStream's own upload validation, which occurs on the backend of the platform when data owners upload files.

You can think of the validation checks in this chapter as a way to identify issues earlier in your reproducible workflow so that data owners do not encounter avoidable upload errors.
:::

<br>

## Data Compilation, Cleaning, and Validation in the Data Rescue Lifecycle

As in the previous chapter, the diagram below situates this chapter within the broader data rescue lifecycle, derived from [Figure 2 in Bledsoe et al. 2025](https://doi.org/10.1098/rspb.2022.0938). This chapter focuses on Data Transfer & Compilation, Data Cleaning & Validation, and Updating metadata as issues emerge.

```{mermaid}
%%| fig-cap: "DataStream example workflow within the broader data rescue lifecycle. This chapter focuses on highlighted boxes: Data transfer & compilation, data cleaning & validation, metadata updates, and preparing outputs for data owner review."
flowchart TD
  S1["Data prioritization<br/><small>Identify high-value or at-risk datasets</small>"]

  S2["Team creation<br/><small>Define roles and responsibilities</small>"]

  S3["Metadata creation<br/><small>Review source files<br/>Create metadata issues table<br/>Create review tables and change log<br/>Draft metadata package</small>"]

  S4["Data transfer & compilation<br/><small>Preserve raw files<br/>Combine source tables<br/>Create inspection table</small>"]

  S5["Data cleaning & validation<br/><small>Clean data<br/>Run QA/QC<br/>Format for DataStream<br/>Run schema validation</small>"]

  U["Update metadata<br/>as issues emerge"]

  R["Prepare package<br/>for data owner review<br/><small>Export final CSV, reports,<br/>validation results, and metadata</small>"]

  S6["Data archiving<br/><small>Data owner uploads finalized<br/>data and metadata to DataStream</small>"]

  S7["Data sharing<br/><small>Archived data are publicly<br/>available for discovery, reuse, and<br/> analysis</small>"]

  S1 --> S2 --> S3 --> S4 --> S5 --> R --> S6 --> S7

  S3 -. "metadata package<br/>is maintained" .-> U
  S4 -. "source-file and<br/>compilation issues" .-> U
  S5 -. "cleaning and validation<br/>decisions" .-> U
  U -. "updated metadata<br/>included in review package" .-> R

  classDef prior fill:#f2f2f2,stroke:#999,color:#666;
  classDef focusData fill:#fff4e6,stroke:#fd7e14,stroke-width:3px,color:#222;
  classDef later fill:#f8f9fa,stroke:#bbb,color:#666;

  class S1,S2,S3 prior;
  class S4,S5,U,R focusData;
  class S6,S7 later;
```

<br>

## Step 1: Load Packages and Create Project Folders

The code in this chapter is meant to be run in your own RStudio session or project environment. You can quickly copy code and paste it into your R session by clicking on the small clipboard in the top right of each code chunk. 

First load the packages that we will need for this chapter.

```{r}
#| message: false
#| echo: true
#| output: false
# Load packages used in this chapter
# The tidyverse packages dplyr, stringr, tibble, readr, and purrr support data cleaning and validation
# while lubridate helps parse and check dates and times
# sf supports spatial validation and mapping
# jsonlite reads the DataStream JSON schema

library(tidyverse)
library(sf)
library(jsonlite)
```

```{r}
#| message: false
#| echo: true
#| output: false
# Create folders for reference files and outputs in your current working directory
# showWarnings = FALSE prevents messages if the folders already exist

dir.create("data_reference", showWarnings = FALSE)
dir.create("outputs", showWarnings = FALSE)
```

<br>

## Step 2: Load the Combined Inspection Table

In the previous chapter, we combined field measurements, lab chemistry results, heavy metal results, and site information into a single table called `dmnp_combined`.

In a real project, this table would usually be created by importing and combining several source files.
To keep this chapter focused on data cleaning, validation, and DataStream formatting, we will begin with the already-combined table.

```{r}
#| message: false
#| echo: true
#| output: true
# Create the table combining source data from the previous chapter
# This represents the output from the previous metadata-preparation chapter

dmnp_combined <- tibble::tribble(
  ~SourceFile, ~DataGroup, ~SiteID, ~SiteName, ~Latitude, ~Longitude,
  ~SampleDate, ~SampleTime, ~SourceCharacteristic, ~ResultRaw, ~UnitRaw,
  ~SampleFraction, ~MethodSpeciation,
  ~DetectionLimitRaw, ~DetectionLimitUnitRaw, ~DetectionLimitTypeRaw,
  ~LaboratoryName, ~AnalyticalMethod,

  "DMNP_Field_2012_2024.xlsx", "Field", "DMNP-ST-01", "North Fork Big Stream",
  46.7351, -60.6412, "2012-07-15", "10:30",
  "DO", "8.4", "mg/L",
  NA_character_, NA_character_,
  NA_character_, NA_character_, NA_character_,
  NA_character_, "YSI Pro Plus",

  "DMNP_Field_2012_2024.xlsx", "Field", "DMNP-ST-02", "South Meadow Creek",
  46.8010, -60.5120, "2014-08-03", "09:15",
  "Temp", "37.8", "deg C",
  NA_character_, NA_character_,
  NA_character_, NA_character_, NA_character_,
  NA_character_, "YSI Pro Plus",

  "DMNP_Field_2012_2024.xlsx", "Field", "DMNP-LK-01", "Clearwater Lake",
  46.9100, -60.4550, "2019-07-21", "13:00",
  "pH", "7.4", "None",
  NA_character_, NA_character_,
  NA_character_, NA_character_, NA_character_,
  NA_character_, "YSI Pro Plus",

  "DMNP_Field_2012_2024.xlsx", "Field", "DMNP-ST-03", "Upper Falls Outlet",
  NA_real_, NA_real_, "2024-08-11", "08:55",
  "SpCond", "125", "uS/cm",
  NA_character_, NA_character_,
  NA_character_, NA_character_, NA_character_,
  NA_character_, "YSI Pro Plus",

  "DMNP_Lab_Chemistry_2012_2024.csv", "Lab chemistry", "DMNP-ST-02", "South meadow creek",
  46.8010, -60.5120, "2018-07-22", "11:45",
  "Nitrate-N", "0.31", "mg/L",
  "Filtered", NA_character_,
  NA_character_, NA_character_, NA_character_,
  "National Laboratory for Environmental Testing", "Colorimetry",

  "DMNP_Lab_Chemistry_2012_2024.csv", "Lab chemistry", "DMNP-LK-01", "Clearwater Lake",
  46.9100, -60.4550, "2019-07-21", "13:05",
  "pH", "7.2", "None",
  NA_character_, NA_character_,
  NA_character_, NA_character_, NA_character_,
  "National Laboratory for Environmental Testing", "SM 4500-H+",

  "DMNP_Lab_Chemistry_2012_2024.csv", "Lab chemistry", "DMNP-ST-03", "Upper Falls Outlet",
  46.8750, -60.5200, "2024-08-11", "09:20",
  "Hardness", "42", "mg/L",
  NA_character_, "From CaCO3",
  NA_character_, NA_character_, NA_character_,
  "National Laboratory for Environmental Testing", "SM 2340B",

  "DMNP_Metals_2016_2024.xlsx", "Heavy metals", "DMNP-LK-01", "Clearwater Lake",
  46.9100, -60.4550, "2020-09-10", "14:20",
  "Pb", "<0.002", "mg/L",
  NA_character_, NA_character_,
  NA_character_, NA_character_, NA_character_,
  "Water Chemistry Lab", NA_character_,

  "DMNP_Metals_2016_2024.xlsx", "Heavy metals", "DMNP-LK-01", "Clearwater Lake",
  46.9100, -60.4550, "2021-09-12", "13:10",
  "Cadmium", "BDL", "ug/L",
  "Dissolved", NA_character_,
  "0.05", "ug/L", "Reporting Limit",
  "Water Chemistry Lab", "EPA 200.8",

  "DMNP_Metals_2016_2024.xlsx", "Heavy metals", "DMNP-ST-02", "South Meadow Creek",
  46.8010, -60.5120, "2022-08-18", "10:05",
  "Copper", "0.004", "mg/L",
  NA_character_, NA_character_,
  NA_character_, NA_character_, NA_character_,
  "Water Chemistry Lab", "EPA 200.8"
)

# Inspect the structure of the combined data table
glimpse(dmnp_combined)
```

This is a combined table from the source data that preserves enough [provenance]{.tooltip-term tabindex="0" data-bs-toggle="tooltip" data-bs-title="Data provenance is the documented history of a data object, including its origin, changes or transformations, and who handled it."} to support cleaning and review.
The next steps will clean selected fields, document decisions, validate scientific plausibility, and format the result for DataStream.

<br>

Let's also import our site reference list from the previous chapter:

```{r}
#| message: false
#| echo: true
#| output: true
# This represents site-level metadata from DMNP_Site_List.xlsx
# The site list acts as the current reference for site names, coordinates, coordinate system, and waterbody type
# We will compare observation records against this reference list

dmnp_site_list <- tibble::tribble(
  ~SiteID, ~SiteName_reference, ~Latitude_reference, ~Longitude_reference,
  ~CoordinateReferenceSystem, ~WaterbodyType,

  "DMNP-ST-01", "North Fork Big Stream", 46.7350, -60.6410,
  "WGS84", "Stream",

  "DMNP-ST-02", "South Meadow Creek", 46.8010, -60.5120,
  "WGS84", "Stream",

  "DMNP-LK-01", "Clearwater Lake", 46.9100, -60.4550,
  "WGS84", "Lake",

  "DMNP-ST-03", "Upper Falls Outlet", 46.8750, -60.5200,
  "WGS84", "Stream"
)

glimpse(dmnp_site_list)
```

<br>

## Step 3: Create a Cleaning Change Log

As introduced in the previous chapter, a **change log** records important changes, decisions, assumptions, and unresolved issues for a dataset.
R is just one way to create and maintain a change log, and the most useful format is a decision that can be made by your team.
The key is that change logs are meant to be read by *humans*, not necessarily machines.
In our example, the change log can be saved as a CSV, spreadsheet, or project document that is updated throughout the workflow.

```{r}
#| message: false
#| echo: true
#| output: true
# Create an empty change log (which we will fill as changes arise)
# Each row will document a data cleaning decision, assumption, or unresolved issue

cleaning_log <- tibble::tibble(
  log_id = integer(),
  processing_area = character(),
  issue_or_decision = character(),
  affected_records = character(),
  action_taken = character(),
  status = character()
)

# View the empty change log
glimpse(cleaning_log)
```

<br>

## Step 4: Clean and Document the Dataset

In the next few code chunks, we will make several demonstrative cleaning decisions.

These steps are not exhaustive, but they are meant to illustrate the types of actions that commonly occur during water quality data rescue, such as:

- Filling in missing coordinates
- Standardizing site names
- Parsing dates and times
- Interpreting source characteristic names
- Parsing censored values (such as `<0.002` and `BDL`)
- Preserving detection-limit information
- Retaining [provenance]{.tooltip-term tabindex="0" data-bs-toggle="tooltip" data-bs-title="Data provenance is the documented history of a data object, including its origin, changes or transformations, and who handled it."} and documenting decisions

### Create a working characteristic lookup table

In the previous chapter, we used working names to interpret source-file terms.
Here, we carry those interpretations forward and begin mapping them to the characteristic names that DataStream uses in their schema.
In the code chunk below, we create a table listing:

1.  The characteristic name used in the source data
2.  The characteristic name used by DataStream that we want to map the source name to
3.  A note briefly justifying the decision

```{r}
#| message: false
#| echo: true
#| output: true
# Create a lookup table for source characteristic names
# CharacteristicName uses DataStream-compatible names for this example
# MethodSpeciation and ResultSampleFraction are retained where they matter for interpretation

characteristic_lookup <- tibble::tribble(
  ~SourceCharacteristic, ~CharacteristicName, ~characteristic_note,

  "DO", "Dissolved oxygen (DO)",
  "Mapped from field abbreviation",

  "Temp", "Temperature, water",
  "Mapped from field abbreviation",

  "SpCond", "Specific conductance",
  "Mapped from field abbreviation",

  "Nitrate-N", "Nitrate",
  "Mapped from lab abbreviation and reported as nitrogen",

  "pH", "pH",
  "Source name already matches a common characteristic name",

  "Pb", "Lead",
  "Mapped from chemical symbol",

  "Cadmium", "Cadmium",
  "Source name already clear",

  "Copper", "Copper",
  "Source name already clear",

  "Hardness", "Total hardness",
  "Mapped from local lab name and reported as CaCO3"
)

# View the structure of the characteristic lookup table
glimpse(characteristic_lookup)
```

### Clean the combined table

The pipe below takes the following steps to clean and update our working data table:

- Create a new table called `dmnp_clean` by joining the combined data table with the site reference list (originally from `DMNP_Site_List.xlsx`) and our lookup table from above for characteristic names (`characteristic_lookup`)
- Change site names and coordinates by mapping to the site reference list
- Use the `lubridate` package to standardize date and time values
- Update time zone values
- Identify and assign censored values to a consistent detection condition

```{r}
#| message: false
#| echo: true
#| output: true
# Clean and enrich the combined table
# This creates a working cleaned dataset before DataStream formatting
# Site names and coordinates are taken from the site list when available
# Dates are parsed with lubridate::ymd so invalid dates can be detected
# Censored values are separated from ordinary numeric result values

dmnp_clean <- dmnp_combined |>
  left_join(dmnp_site_list, by = "SiteID") |>
  left_join(characteristic_lookup, by = "SourceCharacteristic") |>
  mutate(
    # Use the reference site name when available
    SiteNameClean = coalesce(SiteName_reference, SiteName), # dplyr::coalesce is very useful here!

    # Use reference coordinates when source coordinates are missing
    LatitudeClean = coalesce(Latitude, Latitude_reference),
    LongitudeClean = coalesce(Longitude, Longitude_reference),

    # Parse sample dates into Date objects
    SampleDateParsed = lubridate::ymd(SampleDate),

    # Check whether sample times can be parsed as hour-minute values
    SampleTimeParsed = lubridate::hm(SampleTime),

    # Standardize times to HH:MM:SS for later DataStream formatting
    SampleTimeClean = if_else(
      str_detect(SampleTime, "^\\d{2}:\\d{2}$"),
      paste0(SampleTime, ":00"),
      SampleTime
    ),

    # In this fictional example, the data owner confirmed that source times are Atlantic Daylight Time
    # In a real project, do not infer time zones without evidence
    SampleTimeZoneClean = "-04:00",

    # Identify censored values in the source result column
    IsCensored = str_detect(coalesce(ResultRaw, ""), "^<|BDL|Below|ND"),

    # Extract numeric values for ordinary measured results
    ResultValueClean = case_when(
      str_detect(coalesce(ResultRaw, ""), "^<") ~ NA_real_,
      str_detect(coalesce(ResultRaw, ""), "BDL|Below|ND") ~ NA_real_,
      TRUE ~ suppressWarnings(as.numeric(ResultRaw))
    ),

    # Assign a detection condition for censored values
    ResultDetectionConditionClean = if_else(
      IsCensored,
      "Below Detection/Quantification Limit",
      NA_character_
    ),

    # Pull detection-limit values from values like <0.002 or from separate source columns
    DetectionLimitClean = case_when(
      str_detect(coalesce(ResultRaw, ""), "^<") ~
        suppressWarnings(as.numeric(str_remove(ResultRaw, "^<"))),
      !is.na(DetectionLimitRaw) ~
        suppressWarnings(as.numeric(DetectionLimitRaw)),
      TRUE ~ NA_real_
    ),

    # Carry forward detection-limit units and types where they are available
    DetectionLimitUnitClean = case_when(
      IsCensored & !is.na(DetectionLimitUnitRaw) ~ DetectionLimitUnitRaw,
      IsCensored & str_detect(coalesce(ResultRaw, ""), "^<") ~ UnitRaw,
      TRUE ~ NA_character_
    ),

    DetectionLimitTypeClean = case_when(
      IsCensored & !is.na(DetectionLimitTypeRaw) ~ DetectionLimitTypeRaw,
      IsCensored & str_detect(coalesce(ResultRaw, ""), "^<") ~ "Reporting Limit",
      TRUE ~ NA_character_
    )
  )

# View the cleaned data table
glimpse(dmnp_clean)
```

Take a moment to inspect the `dmnp_clean` object.
Notice that the original `Latitude` and `Longitude` columns still contain the values from the source file, including missing values.
The cleaned coordinate fields, `LatitudeClean` and `LongitudeClean`, are the fields used later for DataStream formatting.

Also notice that we did not overwrite the raw source values.
Instead, we created cleaned fields alongside them.
Site names have been standardized in `SiteNameClean`, dates have been parsed in `SampleDateParsed`, times have been reformatted in `SampleTimeClean`, and time zone information has been added in `SampleTimeZoneClean`.
Censored values such as `<0.002` and `BDL` are no longer treated as ordinary numeric results; they are represented with `ResultDetectionConditionClean` and detection-limit fields.
This preserves the original data while making the cleaned values available for validation and DataStream formatting.

### Document cleaning decisions

Now we can document the cleaning actions from above in our change log.

```{r}
#| message: false
#| echo: true
#| output: true
# Document key cleaning decisions in the change log
# This table does not need to capture every line of code
# It should capture decisions that affect interpretation or reproducibility

cleaning_log <- tibble::tribble(
  ~processing_area, ~issue_or_decision, ~affected_records, ~action_taken, ~status,

  "Locations",
  "Some observation records had missing or slightly different coordinates",
  "DMNP-ST-01; DMNP-ST-03",
  "Used DMNP_Site_List.xlsx as the reference source for site names and missing coordinates",
  "Documented decision",

  "Dates and times",
  "Source times were recorded as HH:MM and did not include seconds",
  "All records with SampleTime",
  "Formatted times as HH:MM:SS for DataStream by appending :00",
  "Documented decision",

  "Dates and times",
  "Source files did not include time zone values",
  "All records with SampleTime",
  "Assigned -04:00 after fictional data owner confirmation that source times represent Atlantic Daylight Time",
  "Documented decision",

  "Characteristics",
  "Source files used local abbreviations and names",
  "DO; Temp; SpCond; Pb; Nitrate-N; Hardness",
  "Mapped source terms to working DataStream-compatible characteristic names using a lookup table",
  "Documented decision",

  "Censored values",
  "Some results were reported as censored values rather than exact numeric values",
  "<0.002; BDL",
  "Moved censored results into ResultDetectionCondition and retained detection-limit information where available",
  "Documented decision",

  "Heavy metals",
  "Some heavy metal records were missing sample fraction",
  "Lead; Copper",
  "Retained missing sample fraction as NA and documented limitation rather than guessing dissolved, total, or total recoverable",
  "Unresolved limitation"
) |>
  mutate(log_id = row_number()) |>
  select(log_id, everything())

# View the updated change log
glimpse(cleaning_log)
```

<br>

## Step 5: Scientific Validation and QA/QC

Since most data has been entered by hand at some point, typos and outliers are to be expected!
Outliers may also signal issues with equipment used in the field.
There are several common ways to check the data for outliers and errors.
The ones we will cover here are:

1.  Data bounds
2.  Graphical validation
3.  Spatial validation
4.  Model-based validation

These checks are meant to flag values for review and they should not be used to automatically delete or manipulate raw source data.
Outlier detection is inherently non-trivial with no clear rules, and decisions about removing or modifying values should always be discussed with the project team, a senior researcher, and/or the data owner.

<br>

### Data bounds

Water quality data often has logical or expected limits.
For example, pH values should fall between 0 and 14, and concentrations cannot be negative.

The bounds below are simplified, but demonstrate how we might set up these kinds of checks for similar datasets.

```{r}
#| message: false
#| echo: true
#| output: false
# Create a simple table of logical or expected value bounds
# These bounds are used to flag values for review, not to automatically remove them

bounds_table <- tibble::tribble(
  ~CharacteristicName, ~min_value, ~max_value, ~bounds_note,

  "pH", 0, 14,
  "pH should be between 0 and 14",

  "Temperature, water", -2, 35,
  "Surface water temperature outside this range should be reviewed for this example park",

  "Dissolved oxygen (DO)", 0, 25,
  "Dissolved oxygen should not be negative and very high values should be reviewed",

  "Specific conductance", 0, 5000,
  "Specific conductance should not be negative",

  "Nitrate", 0, 100,
  "Nitrate concentration should not be negative",

  "Total hardness", 0, 1000,
  "Hardness should not be negative",

  "Lead", 0, 100,
  "Lead concentration should not be negative",

  "Cadmium", 0, 100,
  "Cadmium concentration should not be negative",

  "Copper", 0, 100,
  "Copper concentration should not be negative"
)
```

```{r}
#| message: false
#| echo: true
#| output: true
# Check cleaned result values against the bounds table
# Censored values have NA ResultValueClean and are not checked here as exact values

bounds_report <- dmnp_clean |>
  left_join(bounds_table, by = "CharacteristicName") |>
  mutate(row = row_number()) |>
  filter(
    !is.na(ResultValueClean),
    !is.na(min_value),
    ResultValueClean < min_value | ResultValueClean > max_value
  ) |>
  transmute(
    row,
    SourceFile,
    SiteID,
    CharacteristicName,
    ResultValue = ResultValueClean,
    Unit = UnitRaw,
    issue = paste0(
      "Value falls outside expected bounds: ",
      min_value,
      " to ",
      max_value,
      ". ",
      bounds_note
    )
  )

# Inspect the bounds report table
glimpse(bounds_report)
```

In this example, the water temperature value of `37.8 deg C` is flagged for review.
The team should return to the source file, check for transcription or unit errors, and discuss whether the value is plausible for the site and date.

<br>

### Graphical validation

The Decent Mountains National Park example dataset is intentionally small (10 rows of observations), so it is not large enough to demonstrate graphical outlier detection very well.
A plot with only a few observations can help inspect individual records, but it does not show the kinds of patterns we usually look for in larger datasets.

To better demonstrate graphical validation, we will create a larger simulated dataset that follows the same basic structure as the DMNP data.

The goal here is to show how plots can help identify values that are obviously different from the rest of the data or outside expected review bounds.

```{r}
#| message: false
#| echo: true
#| output: false

# Set a seed so the simulated site/date assignments are reproducible

set.seed(123)

# Create a helper function for simulating values for one characteristic
# The ordinary values are generated from normal quantiles so they are approximately normal
# Outlier values are then appended for demonstration

simulate_characteristic <- function(characteristic,
                                    unit,
                                    mean_value,
                                    sd_value,
                                    outlier_values = numeric(0),
                                    n_typical = 40,
                                    digits = 2) {
  
  typical_values <- qnorm(
    ppoints(n_typical),
    mean = mean_value,
    sd = sd_value
  )
  
  values <- round(
    c(sample(typical_values), outlier_values),
    digits = digits
  )
  
  tibble::tibble(
    SiteID = sample(
      c("DMNP-ST-01", "DMNP-ST-02", "DMNP-ST-03", "DMNP-LK-01"),
      size = length(values),
      replace = TRUE
    ),
    SampleDateParsed = sample(
      seq.Date(
        from = as.Date("2012-06-01"),
        to = as.Date("2024-09-30"),
        by = "4 weeks"
      ),
      size = length(values),
      replace = TRUE
    ),
    CharacteristicName = characteristic,
    UnitRaw = unit,
    ResultValueClean = values,
    inserted_for_demo = c(
      rep(FALSE, n_typical),
      rep(TRUE, length(outlier_values))
    )
  )
}
```

```{r}
#| message: false
#| echo: true
#| output: true
# Create a larger fake dataset for graphical QA/QC
# Ordinary values are approximately normally distributed
# Some outliers are intentionally inserted for review

dmnp_graphical_demo <- dplyr::bind_rows(
  simulate_characteristic(
    characteristic = "pH",
    unit = "None",
    mean_value = 7.2,
    sd_value = 0.3,
    outlier_values = c(-0.5, 15.2),
    digits = 2
  ),
  
  simulate_characteristic(
    characteristic = "Temperature, water",
    unit = "deg C",
    mean_value = 14,
    sd_value = 4,
    outlier_values = c(39.5),
    digits = 1
  ),
  
  simulate_characteristic(
    characteristic = "Dissolved oxygen (DO)",
    unit = "mg/L",
    mean_value = 9.5,
    sd_value = 1.2,
    outlier_values = c(-1.2, 26.5),
    digits = 1
  ),
  
  simulate_characteristic(
    characteristic = "Specific conductance",
    unit = "uS/cm",
    mean_value = 125,
    sd_value = 20,
    outlier_values = c(950),
    digits = 0
  ),
  
  simulate_characteristic(
    characteristic = "Nitrate",
    unit = "mg/L",
    mean_value = 0.40,
    sd_value = 0.12,
    outlier_values = c(6.8),
    digits = 2
  ),
  
  simulate_characteristic(
    characteristic = "Total hardness",
    unit = "mg/L",
    mean_value = 45,
    sd_value = 10,
    outlier_values = c(-12, 350),
    digits = 1
  ),
  
  simulate_characteristic(
    characteristic = "Lead",
    unit = "mg/L",
    mean_value = 0.006,
    sd_value = 0.0015,
    outlier_values = c(0.08),
    digits = 4
  ),
  
  simulate_characteristic(
    characteristic = "Cadmium",
    unit = "ug/L",
    mean_value = 0.20,
    sd_value = 0.08,
    outlier_values = c(3.5),
    digits = 2
  ),
  
  simulate_characteristic(
    characteristic = "Copper",
    unit = "mg/L",
    mean_value = 0.006,
    sd_value = 0.002,
    outlier_values = c(0.09),
    digits = 4
  )
)

# Inspect the structure of the simulated dataset
glimpse(dmnp_graphical_demo)
```

Next, we define simple review bounds for the simulated dataset.
These are not universal rules or regulatory thresholds.
They are example screening bounds used to teach graphical outlier detection.

```{r}
#| message: false
#| echo: true
#| output: true
# Create example review bounds for the simulated graphical validation dataset
# These bounds are used to flag values for review, not to automatically delete data

graphical_bounds <- tibble::tribble(
  ~CharacteristicName, ~UnitRaw, ~min_value, ~max_value,
  
  "pH", "None", 0, 14,
  "Temperature, water", "deg C", -2, 35,
  "Dissolved oxygen (DO)", "mg/L", 0, 25,
  "Specific conductance", "uS/cm", 0, 500,
  "Nitrate", "mg/L", 0, 5,
  "Total hardness", "mg/L", 0, 300,
  "Lead", "mg/L", 0, 0.05,
  "Cadmium", "ug/L", 0, 2,
  "Copper", "mg/L", 0, 0.05
)

# View table of graphical bounds
glimpse(graphical_bounds)
```

Now we can compare each simulated result to its review bounds.

```{r}
#| message: false
#| echo: true
#| output: true
# Join the simulated data to the review bounds
# outside_review_bounds identifies values that fall outside the expected range

dmnp_graphical_demo_checked <- dmnp_graphical_demo |>
  left_join(
    graphical_bounds,
    by = c("CharacteristicName", "UnitRaw")
  ) |>
  mutate(
    outside_review_bounds = ResultValueClean < min_value |
      ResultValueClean > max_value,
    plot_label = paste0(CharacteristicName, " (", UnitRaw, ")")
  )

# View the table with simulated data and with values identified outside of bounds
glimpse(dmnp_graphical_demo_checked)
```

Now we can plot all values through time, faceted by characteristic.

```{r}
#| message: false
#| echo: true
#| output: true
# Create a small bounds table for plotting
# This gives ggplot one minimum and maximum line per facet

bounds_for_plot <- dmnp_graphical_demo_checked |>
  distinct(plot_label, min_value, max_value)

# Plot values by date and characteristic
# Facets allow each characteristic to be reviewed on its own scale
# Dashed lines show the example review bounds

dmnp_graphical_demo_checked |>
  ggplot(aes(x = SampleDateParsed, y = ResultValueClean)) +
  geom_hline(
    data = bounds_for_plot,
    aes(yintercept = min_value),
    linetype = "dashed",
    inherit.aes = FALSE
  ) +
  geom_hline(
    data = bounds_for_plot,
    aes(yintercept = max_value),
    linetype = "dashed",
    inherit.aes = FALSE
  ) +
  geom_point(
    aes(shape = outside_review_bounds),
    size = 2
  ) +
  facet_wrap(~ plot_label, scales = "free_y") +
  labs(
    x = "Sample date",
    y = "Result value",
    shape = "Outside review bounds?",
    title = "Graphical outlier review for simulated DMNP water quality data",
    subtitle = "Dashed lines show example review bounds for each characteristic"
  )
```

This plot makes it easy to see which values are unusual relative to values of the same type.

We can also create a table of the records flagged by the graphical review.

```{r}
# Create a report of records that fall outside the graphical review bounds

graphical_outlier_report <- dmnp_graphical_demo_checked |>
  filter(outside_review_bounds) |>
  arrange(CharacteristicName, SampleDateParsed) |>
  transmute(
    SiteID,
    SampleDate = SampleDateParsed,
    CharacteristicName,
    ResultValue = ResultValueClean,
    Unit = UnitRaw,
    expected_range = paste(min_value, "to", max_value, UnitRaw),
    issue = "Value falls outside example review bounds and should be checked"
  )

# View the structure of the report
glimpse(graphical_outlier_report)
```

::: callout-warning
### Do not automatically remove outliers

Graphical validation is a screening step.
A value that looks unusual may be a real environmental observation, a transcription error, a unit error, a method issue, or a result from unusual field conditions.

Before removing or changing values, return to the source files, check field or lab notes, and discuss the record with the project team and/or data owner.
:::

In a real project, you would run this same type of workflow on the actual cleaned dataset rather than on simulated data.
You would also adjust the review bounds, grouping variables, and plots based on the dataset context.
For example, you might review values separately by characteristic, unit, site, season, waterbody type, method, or lab.

<br>

### Validation of spatial data

Coordinates and spatial data should also be checked as part of a cleaning workflow.
In a real project, this might involve plotting points on a map and checking whether they fall inside the expected park boundary or near an expected waterbody.

For this example, we will use the R package `sf` for working with spatial objects to create a simple bounding box around the fictional park and check whether site coordinates fall inside it.

```{r}
#| message: false
#| echo: true
#| output: false
# Create a simple set of unique site coordinates from the cleaned data
# sf::st_as_sf converts the table into a spatial object using longitude and latitude

dmnp_sites_sf <- dmnp_clean |>
  distinct(SiteID, SiteNameClean, LongitudeClean, LatitudeClean) |>
  filter(!is.na(LongitudeClean), !is.na(LatitudeClean)) |>
  st_as_sf(
    coords = c("LongitudeClean", "LatitudeClean"),
    crs = 4326 # coordinate reference system with EPSG code 4326 (This CRS = WGS 84)
  )
```

```{r}
#| message: false
#| echo: true
#| output: false
# Create a simple bounding box for the fictional park using lat/long coordinates
# In a real project, you would use an actual park boundary or expected study area polygon

dmnp_park_bbox <- st_as_sfc(
  st_bbox(
    c(
      xmin = -60.70,
      xmax = -60.40,
      ymin = 46.70,
      ymax = 46.95
    ),
    crs = 4326 # coordinate reference system with EPSG code 4326 (This CRS = WGS 84)
  )
)
```

```{r}
#| message: false
#| echo: true
#| output: true
# Check whether each site falls inside the expected park boundary
# Points outside the boundary would be flagged as FALSE

dmnp_sites_checked <- dmnp_sites_sf |>
  mutate(
    inside_expected_area = lengths(sf::st_within(geometry, dmnp_park_bbox)) > 0
  )

# Plot the site coordinates against the expected study area
# If a point fell outside the rectangle, it would appear outside the bounding box
# and would be marked as FALSE in the legend

ggplot() +
  geom_sf(data = dmnp_park_bbox, fill = NA) +
  geom_sf(
    data = dmnp_sites_checked,
    aes(shape = inside_expected_area),
    size = 2.5
  ) +
  labs(
    title = "DMNP site coordinates for spatial QA/QC",
    subtitle = "Points outside the bounding box would be flagged for review",
    shape = "Inside expected area?"
  )
```
In this example, all site coordinates fall inside the expected study area.
If a coordinate had been entered incorrectly, such as with the wrong sign, swapped latitude and longitude, or a transcription error, the point would appear outside the bounding box and would be marked as `FALSE` in the legend.

The plot is a visual check, but the `inside_expected_area` column is the reproducible check.
That column can be used to create a spatial QA/QC report.
```{r}
#| message: false
#| echo: true
#| output: true
# Check whether each site falls inside the expected park boundary
# lengths(st_within(...)) == 0 identifies points outside the boundary

spatial_report <- dmnp_sites_sf |>
  mutate(
    inside_expected_area = lengths(st_within(geometry, dmnp_park_bbox)) > 0
  ) |>
  filter(!inside_expected_area) |>
  st_drop_geometry() |>
  transmute(
    SiteID,
    SiteName = SiteNameClean,
    issue = "Site coordinate falls outside the expected study area"
  )

spatial_report
```

<br>

### Model-based outlier detection with z-scores

For approximately normally distributed data, a simple model-based approach is to calculate [z-scores]{.tooltip-term tabindex="0" data-bs-toggle="tooltip" data-bs-title="A z-score tells you how many standard deviations a data point is from the mean, helping you spot values that are typical or unusual."} within each characteristic.
A common rule of thumb is to flag values with z-scores less than `-3` or greater than `3`, though the type of dataset and its ecological context is likely to influence this interpretation.

Note that this is only a screening method, and tt should not be used blindly for small datasets, skewed data, censored data, or data with strong seasonal or site effects.

We will use the simulated data from above to first check that our data follow (approximately) normal distributions.

```{r}
#| message: false
#| echo: true
#| output: true
# Check whether values for each characteristic are approximately normally distributed
# The Shapiro-Wilk test is used here as a simple screening check
# Small p-values suggest that the values may not be normally distributed

normality_review <- dmnp_graphical_demo |>
  filter(
    !is.na(ResultValueClean),
    inserted_for_demo == FALSE
  ) |>
  group_by(CharacteristicName) |>
  summarise(
    n_values = n(),
    mean_value = mean(ResultValueClean, na.rm = TRUE),
    sd_value = sd(ResultValueClean, na.rm = TRUE),
    shapiro_p_value = if (n_values >= 3) {
      shapiro.test(ResultValueClean)$p.value
    } else {
      NA_real_
    },
    normality_note = case_when(
      is.na(shapiro_p_value) ~ "Too few values to assess normality",
      shapiro_p_value < 0.05 ~ "Values may not be normally distributed; use z-scores cautiously",
      TRUE ~ "No strong evidence against normality from this screening check"
    ),
    .groups = "drop"
  )

normality_review
```
Our variables do not appear to violate normality, so we can proceed with using z-scores (but know that other options are available if data does not adhere to a normal distribution).

```{r}
#| message: false
#| echo: true
#| output: true
# Calculate simple z-scores within each characteristic on the simulated data
# This is most useful when there are enough comparable records within a group
# The small DMNP example may not have enough values for this to be very informative

zscore_report <- dmnp_graphical_demo |>
  filter(!is.na(ResultValueClean)) |>
  group_by(CharacteristicName) |>
  mutate(
    n_values = n(),
    mean_value = mean(ResultValueClean, na.rm = TRUE),
    sd_value = sd(ResultValueClean, na.rm = TRUE),
    z_score = if_else(
      n_values >= 3 & !is.na(sd_value) & sd_value > 0,
      (ResultValueClean - mean_value) / sd_value,
      NA_real_
    )
  ) |>
  ungroup() |>
  filter(!is.na(z_score), abs(z_score) > 3) |>
  transmute(
    SiteID,
    CharacteristicName,
    ResultValue = ResultValueClean,
    z_score,
    issue = "Absolute z-score is greater than 3; review as possible outlier"
  )

# Check which values have absolute z-scores greater than 3
glimpse(zscore_report)
```

It looks like we have a number of outliers from our simulated dataset that we will want to flag. In many real water quality datasets, you could refine this approach by grouping values by characteristic, unit, site type, season, or monitoring program.
Always discuss suspected outliers before removing or changing data.

<br>

### Summarize scientific QA/QC

We can now summarize our scientific validation efforts in a saved report.
Because the graphical validation section uses simulated teaching data, those records are included in the report with the label `Graphical validation demo`.

```{r}
#| message: false
#| echo: true
#| output: true
# Combine scientific validation reports into one QA/QC report
# Each report is converted to a shared format before combining
# The graphical validation report comes from simulated teaching data, so we label it clearly

scientific_qaqc_report <- bind_rows(
  bounds_report |>
    transmute(
      check_type = "Data bounds",
      record = paste(SiteID, CharacteristicName, sep = " | "),
      value = as.character(ResultValue),
      issue
    ),

  zscore_report |>
    transmute(
      check_type = "Z-score outlier screen",
      record = paste(SiteID, CharacteristicName, sep = " | "),
      value = as.character(ResultValue),
      issue
    ),

  graphical_outlier_report |>
    transmute(
      check_type = "Graphical validation demo",
      record = paste(SiteID, CharacteristicName, sep = " | "),
      value = paste(ResultValue, Unit),
      issue = paste(issue, "Expected range:", expected_range)
    ),

  spatial_report |>
    transmute(
      check_type = "Spatial validation",
      record = paste(SiteID, SiteName, sep = " | "),
      value = NA_character_,
      issue
    )
)

# View the report
glimpse(scientific_qaqc_report)
```

```{r}
#| message: false
#| echo: true
#| output: false
# Save the scientific QA/QC report
# This report can be shared with team members for review

if (nrow(scientific_qaqc_report) > 0) {
  readr::write_csv(
    scientific_qaqc_report,
    "outputs/dmnp_scientific_qaqc_report.csv"
  )
}
```

<br>

## Step 7: Format the Dataset for DataStream

After cleaning and scientific validation, we can create a DataStream-formatted table.

This step translates our cleaned working table, `dmnp_clean`, into DataStream's expected field names and controlled vocabulary values.
This is one of the most important steps in the workflow because it is where project-specific column names, units, methods, and metadata are converted into a structure that can be uploaded to DataStream.

In the code below, we are doing several things at once:

- Selecting the columns that will be included in the DataStream export
- Renaming cleaned project-specific columns to DataStream field names
- Adding dataset-level information that applies to every row
- Recoding source values to DataStream controlled vocabulary values
- Carrying forward cleaned dates, times, locations, methods, units, and result values
- Representing censored values using detection-condition and detection-limit fields
- Preserving important unresolved issues in `ResultComment`

The function `transmute()` is useful here because it creates a new table that contains only the columns listed inside the function.
This helps ensure that extra working columns from the cleaning process are not accidentally included in the final export.

```{r}
#| message: false
#| echo: true
#| output: true
# Create a DataStream-formatted export table
# dmnp_clean is the cleaned working table
# datastream_export is the table we are preparing for DataStream upload

datastream_export <- dmnp_clean |>
  transmute(
    # Dataset-level information
    # This value is repeated on every row of the upload file

    DatasetName = "Decent Mountains National Park Freshwater Monitoring",

    # Monitoring location information
    # These columns use the cleaned site names and coordinates created earlier

    MonitoringLocationID = SiteID,
    MonitoringLocationName = SiteNameClean,
    MonitoringLocationLatitude = LatitudeClean,
    MonitoringLocationLongitude = LongitudeClean,
    MonitoringLocationHorizontalCoordinateReferenceSystem =
      CoordinateReferenceSystem,

    # Recode source waterbody types to DataStream location types
    # Unknown or unresolved values are left blank rather than guessed

    MonitoringLocationType = case_when(
      WaterbodyType == "Stream" ~ "River/Stream",
      WaterbodyType == "Lake" ~ "Lake/Pond",
      TRUE ~ NA_character_
    ),

    # Activity information
    # Field measurements and lab-analyzed samples use different activity types

    ActivityType = case_when(
      DataGroup == "Field" ~ "Field Msr/Obs",
      TRUE ~ "Sample-Routine"
    ),

    # All observations in this example are from surface water

    ActivityMediaName = "Surface Water",

    # Date and time information
    # These fields use the cleaned date, time, and time zone values

    ActivityStartDate = as.character(SampleDateParsed),
    ActivityStartTime = SampleTimeClean,
    ActivityStartTimeZone = SampleTimeZoneClean,

    # Characteristic and result metadata
    # These fields describe what was measured and how the result should be interpreted

    CharacteristicName = CharacteristicName,
    MethodSpeciation = MethodSpeciation,
    ResultSampleFraction = SampleFraction,

    # Result values
    # Exact numeric results go in ResultValue
    # Censored results have NA here and are described in detection-condition fields below

    ResultValue = ResultValueClean,
    ResultUnit = UnitRaw,
    ResultValueType = "Actual",

    # Detection-limit information
    # These fields are used for censored values such as <0.002 or BDL

    ResultDetectionCondition = ResultDetectionConditionClean,
    ResultDetectionQuantitationLimitMeasure = DetectionLimitClean,
    ResultDetectionQuantitationLimitUnit = DetectionLimitUnitClean,
    ResultDetectionQuantitationLimitType = DetectionLimitTypeClean,

    # Result status
    # This example assumes final results, but real projects should confirm this

    ResultStatusID = "Final",

    # Method and laboratory information
    # Field records use the field instrument name
    # Lab records carry forward the analytical method and lab name from the source data

    ResultAnalyticalMethodName = case_when(
      DataGroup == "Field" ~ "YSI Pro Plus multiparameter meter",
      TRUE ~ AnalyticalMethod
    ),

    LaboratoryName = if_else(
      DataGroup == "Field",
      NA_character_,
      LaboratoryName
    ),

    # Record-level comments
    # These comments preserve important unresolved issues without guessing

    ResultComment = case_when(
      SourceCharacteristic == "Pb" & is.na(SampleFraction) ~
        "Sample fraction missing in source file; result should not be assumed dissolved, total, or total recoverable",

      SourceCharacteristic == "Copper" & is.na(SampleFraction) ~
        "Sample fraction missing in source file; result should not be assumed dissolved, total, or total recoverable",

      TRUE ~ NA_character_
    )
  )

# View the formatted table
# The table now uses DataStream field names, but still needs schema validation

glimpse(datastream_export)
```

```{r}
# Write the DataStream-formatted table to CSV
# This is the file that would be shared with the data owner for review and upload

readr::write_csv(
  datastream_export,
  "outputs/dmnp_datastream_export.csv"
)
```

At this point, `datastream_export` has been reshaped into a DataStream-style table.
However, formatting the table does not guarantee that every value is valid.
The next step is to check the table against DataStream's JSON schema.



<br>

## Step 8: Check Adherence to the DataStream Schema

The code block below downloads the current DataStream JSON schema, extracts key schema rules, and runs a set of pre-upload checks on `datastream_export`. It is primarily intended as a reference for checking if your dataset adheres to DataStream's expected formatting.

You do not need to understand every line of this code chunk, and it is long!
In your own project, you can treat this as a validation utility: run it, inspect the report, fix issues upstream in your cleaning workflow, and run it again.

```{r}
#| message: false
#| echo: true
#| output: true

# Download, read, and use the DataStream JSON schema to validate a formatted dataset
# This block creates one validation report and saves it to outputs/datastream_validation_report.csv

# Download the current DataStream schema in JSON format
download.file(
  url = "https://datastream.org/schema",
  destfile = "data_reference/datastream_schema.json",
  mode = "wb"
)

# Read the schema into R
datastream_schema <- jsonlite::read_json(
  path = "data_reference/datastream_schema.json",
  simplifyVector = FALSE
)

# Record the schema version
schema_version <- datastream_schema[["$comment"]]

# Helper operator for replacing NULL with a fallback value
`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

# Extract the main schema object
schema_core <- datastream_schema$allOf[[1]]
schema_properties <- schema_core$properties
schema_fields <- names(schema_properties)
required_fields <- schema_core$required %||% character(0)
schema_dependencies <- schema_core$dependencies %||% list()

# Helper functions for extracting schema details
get_chr <- function(x) {
  if (is.null(x)) {
    NA_character_
  } else {
    paste(unlist(x), collapse = ", ")
  }
}

get_num <- function(x) {
  if (is.null(x)) {
    NA_real_
  } else {
    as.numeric(x[[1]])
  }
}

# Create a compact schema summary
schema_field_summary <- purrr::imap_dfr(
  schema_properties,
  function(field_schema, field_name) {
    tibble::tibble(
      field = field_name,
      type = get_chr(field_schema$type),
      format = get_chr(field_schema$format),
      minimum = get_num(field_schema$minimum),
      maximum = get_num(field_schema$maximum),
      min_length = get_num(field_schema$minLength),
      max_length = get_num(field_schema$maxLength),
      has_allowed_values = !is.null(field_schema$enum)
    )
  }
)

# Extract allowed values from enum fields
allowed_values <- purrr::imap_dfr(
  schema_properties,
  function(field_schema, field_name) {
    if (is.null(field_schema$enum)) {
      return(tibble::tibble())
    }

    tibble::tibble(
      field = field_name,
      allowed_value = as.character(unlist(field_schema$enum))
    )
  }
)

# Convert schema dependencies to a table
dependency_table <- purrr::imap_dfr(
  schema_dependencies,
  function(dependent_fields, trigger_field) {
    tibble::tibble(
      trigger_field = trigger_field,
      dependent_field = as.character(unlist(dependent_fields))
    )
  }
)

# Helper for blank values
is_blank <- function(x) {
  is.na(x) | stringr::str_trim(as.character(x)) == ""
}

# Create a consistent empty validation table
empty_issue <- function() {
  tibble::tibble(
    row = integer(),
    field = character(),
    value = character(),
    issue = character()
  )
}

# Helper for making validation issue tables
make_issue <- function(data, rows, field, issue) {
  if (length(rows) == 0) {
    return(empty_issue())
  }

  values <- if (field %in% names(data)) {
    as.character(data[[field]][rows])
  } else {
    rep(NA_character_, length(rows))
  }

  tibble::tibble(
    row = rows,
    field = field,
    value = values,
    issue = issue
  )
}

# Check for columns not defined in the schema
check_extra_columns <- function(data, schema_fields) {
  extra_cols <- setdiff(names(data), schema_fields)

  if (length(extra_cols) == 0) {
    return(empty_issue())
  }

  tibble::tibble(
    row = NA_integer_,
    field = extra_cols,
    value = NA_character_,
    issue = "Column is not defined in the DataStream schema"
  )
}

# Check required fields
check_required_fields <- function(data, required_fields) {
  missing_required <- setdiff(required_fields, names(data))

  missing_issues <- if (length(missing_required) > 0) {
    tibble::tibble(
      row = NA_integer_,
      field = missing_required,
      value = NA_character_,
      issue = "Required field is missing from the dataset"
    )
  } else {
    empty_issue()
  }

  present_required <- intersect(required_fields, names(data))

  blank_issues <- if (length(present_required) > 0) {
    purrr::map_dfr(present_required, function(field) {
      bad_rows <- which(is_blank(data[[field]]))

      make_issue(
        data = data,
        rows = bad_rows,
        field = field,
        issue = "Required field is blank"
      )
    })
  } else {
    empty_issue()
  }

  dplyr::bind_rows(missing_issues, blank_issues)
}

# Check controlled vocabulary fields
check_allowed_values <- function(data, allowed_values) {
  fields_to_check <- intersect(
    unique(allowed_values$field),
    names(data)
  )

  if (length(fields_to_check) == 0) {
    return(empty_issue())
  }

  purrr::map_dfr(fields_to_check, function(field) {
    allowed <- allowed_values |>
      dplyr::filter(.data$field == .env$field) |>
      dplyr::pull(allowed_value)

    bad_rows <- which(
      !is_blank(data[[field]]) &
        !(as.character(data[[field]]) %in% allowed)
    )

    make_issue(
      data = data,
      rows = bad_rows,
      field = field,
      issue = "Value is not listed as an allowed value in the DataStream schema"
    )
  })
}

# Check numeric fields and schema-defined ranges
check_numeric_ranges <- function(data, schema_field_summary) {
  numeric_fields <- schema_field_summary |>
    dplyr::filter(type %in% c("number", "integer")) |>
    dplyr::filter(field %in% names(data))

  if (nrow(numeric_fields) == 0) {
    return(empty_issue())
  }

  purrr::map_dfr(seq_len(nrow(numeric_fields)), function(i) {
    field <- numeric_fields$field[i]
    min_value <- numeric_fields$minimum[i]
    max_value <- numeric_fields$maximum[i]

    x <- suppressWarnings(as.numeric(data[[field]]))

    bad_rows <- which(
      !is_blank(data[[field]]) &
        (
          is.na(x) |
            (!is.na(min_value) & x < min_value) |
            (!is.na(max_value) & x > max_value)
        )
    )

    make_issue(
      data = data,
      rows = bad_rows,
      field = field,
      issue = "Value is not numeric or falls outside the range defined in the DataStream schema"
    )
  })
}

# Check dates, times, and time zones
check_date_time_formats <- function(data, schema_field_summary) {
  date_fields <- schema_field_summary |>
    dplyr::filter(format == "date") |>
    dplyr::pull(field) |>
    intersect(names(data))

  date_issues <- if (length(date_fields) > 0) {
    purrr::map_dfr(date_fields, function(field) {
      bad_rows <- which(
        !is_blank(data[[field]]) &
          is.na(lubridate::ymd(data[[field]]))
      )

      make_issue(
        data = data,
        rows = bad_rows,
        field = field,
        issue = "Date should be parseable as YYYY-MM-DD"
      )
    })
  } else {
    empty_issue()
  }

  time_fields <- intersect(
    c("ActivityStartTime", "ActivityEndTime", "AnalysisStartTime"),
    names(data)
  )

  time_pattern <- "^(2[0-3]|[01][0-9]):[0-5][0-9](:[0-5][0-9](\\.[0-9]{1,3})?)?$"

  time_issues <- if (length(time_fields) > 0) {
    purrr::map_dfr(time_fields, function(field) {
      bad_rows <- which(
        !is_blank(data[[field]]) &
          !stringr::str_detect(as.character(data[[field]]), time_pattern)
      )

      make_issue(
        data = data,
        rows = bad_rows,
        field = field,
        issue = "Time should use HH:MM, HH:MM:SS, or HH:MM:SS.sss format"
      )
    })
  } else {
    empty_issue()
  }

  timezone_fields <- intersect(
    c("ActivityStartTimeZone", "ActivityEndTimeZone", "AnalysisStartTimeZone"),
    names(data)
  )

  timezone_pattern <- "^(Z|[+-](0[0-9]|1[0-9]|2[0-4]):[0-5][0-9])$"

  timezone_issues <- if (length(timezone_fields) > 0) {
    purrr::map_dfr(timezone_fields, function(field) {
      bad_rows <- which(
        !is_blank(data[[field]]) &
          !stringr::str_detect(as.character(data[[field]]), timezone_pattern)
      )

      make_issue(
        data = data,
        rows = bad_rows,
        field = field,
        issue = "Time zone should use Z or +/-HH:MM format"
      )
    })
  } else {
    empty_issue()
  }

  dplyr::bind_rows(date_issues, time_issues, timezone_issues)
}

# Check simple dependency rules
check_dependencies <- function(data, dependency_table) {
  if (nrow(dependency_table) == 0) {
    return(empty_issue())
  }

  purrr::map_dfr(seq_len(nrow(dependency_table)), function(i) {
    trigger_field <- dependency_table$trigger_field[i]
    dependent_field <- dependency_table$dependent_field[i]

    if (!trigger_field %in% names(data)) {
      return(empty_issue())
    }

    trigger_rows <- which(!is_blank(data[[trigger_field]]))

    if (length(trigger_rows) == 0) {
      return(empty_issue())
    }

    if (!dependent_field %in% names(data)) {
      return(
        tibble::tibble(
          row = trigger_rows,
          field = dependent_field,
          value = NA_character_,
          issue = paste0(
            "Field is required when ",
            trigger_field,
            " is populated, but this column is missing"
          )
        )
      )
    }

    bad_rows <- trigger_rows[
      is_blank(data[[dependent_field]][trigger_rows])
    ]

    make_issue(
      data = data,
      rows = bad_rows,
      field = dependent_field,
      issue = paste0(
        "Required when ",
        trigger_field,
        " is populated"
      )
    )
  })
}

# Check ResultValue and ResultDetectionCondition relationship
check_result_value_detection_condition <- function(data) {
  required_fields <- c("ResultValue", "ResultDetectionCondition")

  if (!all(required_fields %in% names(data))) {
    return(empty_issue())
  }

  both_blank <- which(
    is_blank(data$ResultValue) &
      is_blank(data$ResultDetectionCondition)
  )

  both_populated <- which(
    !is_blank(data$ResultValue) &
      !is_blank(data$ResultDetectionCondition)
  )

  dplyr::bind_rows(
    make_issue(
      data = data,
      rows = both_blank,
      field = "ResultValue",
      issue = "Either ResultValue or ResultDetectionCondition should be populated"
    ),
    make_issue(
      data = data,
      rows = both_populated,
      field = "ResultDetectionCondition",
      issue = "Do not populate ResultDetectionCondition when ResultValue is also populated"
    )
  )
}

# Check lab method requirements for routine samples
check_lab_method_fields <- function(data) {
  if (!"ActivityType" %in% names(data)) {
    return(empty_issue())
  }

  sample_rows <- which(
    stringr::str_detect(
      dplyr::coalesce(data$ActivityType, ""),
      "^Sample-"
    )
  )

  if (length(sample_rows) == 0) {
    return(empty_issue())
  }

  get_field <- function(field) {
    if (field %in% names(data)) {
      as.character(data[[field]])
    } else {
      rep(NA_character_, nrow(data))
    }
  }

  laboratory_name <- get_field("LaboratoryName")
  method_name <- get_field("ResultAnalyticalMethodName")
  method_id <- get_field("ResultAnalyticalMethodID")
  method_context <- get_field("ResultAnalyticalMethodContext")

  lab_missing <- sample_rows[
    is_blank(laboratory_name[sample_rows])
  ]

  method_missing <- sample_rows[
    is_blank(method_name[sample_rows]) &
      (
        is_blank(method_id[sample_rows]) |
          is_blank(method_context[sample_rows])
      )
  ]

  dplyr::bind_rows(
    make_issue(
      data = data,
      rows = lab_missing,
      field = "LaboratoryName",
      issue = "LaboratoryName should be populated for lab-analyzed samples"
    ),
    make_issue(
      data = data,
      rows = method_missing,
      field = "ResultAnalyticalMethodName",
      issue = "Analytical method information should be populated for lab-analyzed samples"
    )
  )
}

# Run all checks and combine them into one report
validation_report <- dplyr::bind_rows(
  check_extra_columns(datastream_export, schema_fields),
  check_required_fields(datastream_export, required_fields),
  check_allowed_values(datastream_export, allowed_values),
  check_numeric_ranges(datastream_export, schema_field_summary),
  check_date_time_formats(datastream_export, schema_field_summary),
  check_dependencies(datastream_export, dependency_table),
  check_result_value_detection_condition(datastream_export),
  check_lab_method_fields(datastream_export)
)

# Arrange validation issues when any are found
if (nrow(validation_report) > 0) {
  validation_report <- validation_report |>
    dplyr::arrange(row, field)
} else {
  validation_report <- empty_issue()
}

# Save the validation report
readr::write_csv(
  validation_report,
  "outputs/datastream_validation_report.csv"
)

# Print the schema version and validation report
schema_version
validation_report

# Message for the user
if (nrow(validation_report) == 0) {
  message("No issues found by the pre-upload schema checks")
} else {
  message(
    "Pre-upload schema checks found issues. ",
    "Review outputs/datastream_validation_report.csv, fix issues upstream, and rerun validation."
  )
}
```

The validation report found one issue: `ResultAnalyticalMethodName` is missing for one lab-analyzed sample.

In this case, the issue is not with the result value itself.
The issue is that the data record does not include enough analytical method information for a lab-analyzed result.
Because analytical methods affect how a result should be interpreted, we should not fill this field by guessing.

Instead, we should flag the record for review, return to the source files, and ask the data owner to confirm the appropriate analytical method if it cannot be found in the available documentation.

```{r}
#| message: false
#| echo: true
#| output: true
# Identify records flagged for missing analytical method information
# These records should be reviewed against source files or confirmed with the data owner

method_issue_rows <- validation_report |>
  filter(field == "ResultAnalyticalMethodName") |>
  pull(row) |>
  unique()

method_review <- datastream_export |>
  mutate(row = row_number()) |>
  filter(row %in% method_issue_rows) |>
  select(
    any_of(c(
      "row",
      "MonitoringLocationID",
      "MonitoringLocationName",
      "ActivityStartDate",
      "CharacteristicName",
      "ResultValue",
      "ResultUnit",
      "LaboratoryName",
      "ResultAnalyticalMethodName",
      "ResultComment"
    ))
  )

method_review
```

This issue should be added to the metadata change log or unresolved issues list.
If the analytical method is later confirmed, the correction should be made upstream in the cleaning or formatting workflow and the validation checks can be run again.

<br>

## Step 9: Review and Export Final Outputs

The final outputs from this workflow are:

- A cleaned working table
- A cleaning change log
- A scientific QA/QC report
- A DataStream-formatted export table
- A DataStream schema validation report

```{r}
#| message: false
#| echo: true
#| output: false
# Save the cleaned working table
# This preserves the cleaned version of the project-specific dataset before final DataStream formatting

readr::write_csv(
  dmnp_clean,
  "outputs/dmnp_clean_working_table.csv"
)

# Save the cleaning change log
# This documents key processing decisions, assumptions, and unresolved issues

readr::write_csv(
  cleaning_log,
  "outputs/dmnp_cleaning_log.csv"
)

# Save the scientific QA/QC report
# This includes records flagged during data bounds, outlier, graphical, or spatial checks

readr::write_csv(
  scientific_qaqc_report,
  "outputs/dmnp_scientific_qaqc_report.csv"
)

# Save the final DataStream-formatted export
# This is the main CSV that would be shared with the data owner for review and upload

readr::write_csv(
  datastream_export,
  "outputs/dmnp_datastream_export.csv"
)

# Save the DataStream schema validation report
# This records any remaining issues found by the pre-upload schema checks

readr::write_csv(
  validation_report,
  "outputs/datastream_validation_report.csv"
)
```

<br>

## Step 10: Update Metadata Documentation

At the beginning of this chapter, we noted that updates to metadata are often continuous throughout a project.
Now that we have performed cleaning, validation, and formatting of the dataset, we should update the metadata documentation from the previous chapter.

For the Decent Mountains National Park example, the metadata should now document that:

- The site list was used as the reference source for site names and missing coordinates
- Sample times were reformatted from `HH:MM` to `HH:MM:SS`
- The time zone was assigned as `-04:00` after data owner confirmation
- Censored values such as `<0.002` and `BDL` were represented using detection-condition and detection-limit fields
- Missing sample fraction remains unresolved for some heavy metal records
- A high water temperature value was flagged for review during scientific QA/QC
- The hardness result was reformatted after schema validation because `From CaCO3` was not an allowed `MethodSpeciation` value

```{r}
# Add metadata updates that emerged during data cleaning and validation
# This table follows the same structure as the metadata change log from the previous chapter

# Identify any rows flagged for missing analytical method information
method_issue_rows <- validation_report |>
  filter(field == "ResultAnalyticalMethodName") |>
  pull(row) |>
  unique()

method_issue_records <- if (length(method_issue_rows) > 0) {
  paste(method_issue_rows, collapse = "; ")
} else {
  "No records flagged"
}

metadata_change_log_updates <- tibble::tribble(
  ~metadata_area, ~issue_or_decision, ~affected_records, ~evidence, ~action_taken, ~status,

  "Locations",
  "Used the site list as the reference source for site names and missing coordinates",
  "DMNP-ST-01; DMNP-ST-03",
  "Cleaning workflow used SiteName_reference, Latitude_reference, and Longitude_reference from DMNP_Site_List.xlsx",
  "Update Data Processing metadata to state that site metadata were standardized using the site list",
  "Documented decision",

  "Dates and times",
  "Sample times were reformatted from HH:MM to HH:MM:SS and assigned time zone -04:00 after data owner confirmation",
  "All records with SampleTime",
  "Cleaning workflow created SampleTimeClean and SampleTimeZoneClean",
  "Update Data Processing metadata and note that seconds were added for formatting rather than measured in the source data",
  "Documented decision",

  "Censored values",
  "Censored values were represented using ResultDetectionCondition and detection-limit fields where possible",
  "<0.002; BDL",
  "Cleaning workflow created ResultDetectionConditionClean, DetectionLimitClean, DetectionLimitUnitClean, and DetectionLimitTypeClean",
  "Update Data Processing metadata to describe how censored values were handled",
  "Documented decision",

  "Heavy metals",
  "Sample fraction remains missing for selected lead and copper records",
  "Lead; Copper",
  "ResultComment was used to document that these results should not be assumed dissolved, total, or total recoverable",
  "Keep this as an unresolved limitation in the metadata package and data owner questions",
  "Unresolved limitation",

  "Scientific QA/QC",
  "A high water temperature value was flagged for review",
  "Temperature, water",
  "Data bounds check flagged 37.8 deg C as outside the example review bounds",
  "Add to the QA/QC summary and review the source file before changing or excluding the value",
  "Needs review",

  "Lab methods",
  "One lab-analyzed result is missing analytical method information",
  method_issue_records,
  "Pre-upload validation flagged missing ResultAnalyticalMethodName for a lab-analyzed sample",
  "Review source method files or confirm the analytical method with the data owner before updating the DataStream export",
  "Needs review"
) |>
  mutate(log_id = row_number())

glimpse(metadata_change_log_updates)
```

These entries should be incorporated into the metadata package from the previous chapter, particularly the change log, unresolved issues list, and draft **Data Processing** field.
The goal is for the metadata to continually reflect what happened during data cleaning, scientific QA/QC, and schema validation.

## Review: What These Checks Do and Do Not Do

| Question | Addressed in this chapter? |
|----|----|
| Were source files combined in a reproducible way? | Yes |
| Were cleaning decisions documented? | Yes |
| Were censored values parsed and documented? | Yes |
| Were missing coordinates reviewed against a site list? | Yes |
| Were scientifically suspicious values flagged? | Yes |
| Were coordinates checked spatially? | Yes |
| Were columns and values checked against the DataStream schema? | Yes |
| Does this replace DataStream's upload validator? | No |
| Does this prove every value is valid? | No |
| Should flagged outliers be automatically removed? | No |

Scientific validation and schema validation are complementary.
A value can pass the DataStream schema and still be scientifically suspicious.
Likewise, a scientifically plausible value may still fail schema validation if it uses an invalid unit, missing required field, or non-standard controlled vocabulary value.

When validation reports identify issues, return to the cleaning workflow, update the code, document the decision in the change log, and rerun the checks.
The goal is to create a dataset that is reproducible, well-documented, scientifically defensible, and ready for data owner review.

<br>

## R packages used in this chapter

Pebesma, E. (2018). Simple Features for R: Standardized Support for Spatial Vector Data. *The R Journal* 10 (1), 439-446,
<https://doi.org/10.32614/RJ-2018-009>.

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, Max, Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., Takahashi, K., Vaughan, D., Wilke, C., Woo, K., & Yutani, H. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, *4*(43), 1686. <https://doi.org/10.21105/joss.01686>.

## References in this chapter

Bledsoe, E. K., Burant, J. B., Higino, G. T., Roche, D. G., Binning, S. A., Finlay, K., Pither, J., Pollock, L. S., Sunday, J. M., & Srivastava, D. S. (2022). Data rescue: Saving environmental data from extinction. Proceedings of the Royal Society B: Biological Sciences, 289(1979), Article 20220938. <https://doi.org/10.1098/rspb.2022.0938>.

DataStream Initiative. 2021. The DataStream Learning Centre. <https://datastream.org/en-ca/documentation>.
